---
title: "Two species SIR model with vaccination"
author: Ottar N. Bjørnstad
output: html_document
---


This Rmarkdown is of a 2sp-SIR-with-vaccination model. 
The vectored equations for the flow of hosts between **S**usceptible, **I**nfectious and **R**ecovered
 compartments are:
 
$\begin{aligned}
    \frac{d \vec{S}}{dt} &= \underbrace{\mu \vec{N} (1-\vec{p})}_{\mbox{recruitment}} - \underbrace{\phi \vec{S}}_{\mbox{infection}} - \underbrace{\mu \vec{S}}_{\mbox{death}} \label{eq:sirs}\\
    &\underbrace{\phi = \beta \times (\vec{I} / \vec{N})}_{\mbox{force of infection}}\\
     \frac{d\vec{I}}{dt} &= \underbrace{\phi \vec{S}}_{\mbox{infection}} - \underbrace{\gamma \vec{I}}_{\mbox{recovery}} - \underbrace{\mu \vec{I}}_{\mbox{death}}  \label{eq:siri}\\
     \frac{d\vec{R}}{dt} &= \underbrace{\gamma \vec{I}}_{\mbox{recovery}} - \underbrace{\mu \vec{R}}_{\mbox{death}} \label{eq:sirr}
\end{aligned}$

The transmission matrix $\beta$ is 2-by-2 where the diagonal is within species transmission and the off-diagonals are (possibly assymetric) cross-species transmission. The length-2 vectors, $\mu, p$ and $\gamma$ represents species-specific birth rates, vaccination-at-birth probabilities and recovery rates, respectively.

```{r, echo=FALSE}
using<-function(...) {
    libs<-unlist(list(...))
    req<-unlist(lapply(libs,require,character.only=TRUE))
    need<-libs[req==FALSE]
    if(length(need)>0){ 
        install.packages(need)
        lapply(need,require,character.only=TRUE)
    }
}

using("deSolve")
```

```{r}
  #Gradient function;
  twosp=function(t, x, parms){
    S=x[1:2]
    I=x[3:4]
    R=x[5:6]

    beta=parms$beta
    mu=parms$mu
    gamma=parms$gamma
    N=parms$N
    p=parms$p
    phi = (beta %*% I)/N
    dS = (1-p)*mu * N  - phi * S  - mu * S 
    dI = phi * S - (mu + gamma) * I
    dR = gamma * I - mu * R + p * R
    res=c(dS, dI, dR)
    list(res)
  }
```
Define parameters and set up for solver:
```{r}
  times  = seq(0, 20, by=1/10)
  parms  = list(mu = c(0.2, 0.1), N = c(1,1), 
  beta =  matrix(c(3, 2, 
                           1, 1), ncol=2, byrow=TRUE), 
  gamma = c(1,1),
  p=c(.1, 0))
```
Run solver:
```{r}
  start = c(S=c(0.999, 0.999), I=c(0.001, 0.001), R = c(0,0))
 out=ode(y=start,  times=times, func=twosp, parms=parms)
  out=as.data.frame(out)
  matplot(out[,4:5])
```

Now set up for R0 next gen calc:

```{r}
nextgenR0=function(Istates, Flist, Vlist, params, dfe){
paras = as.list(c(dfe, params)) 

k=0
vl=fl=list(NULL)
for(i in 1:length(Istates)){
assign(paste("f", i, sep = "."), lapply(lapply(Flist,deriv, Istates[i]), eval, paras))
assign(paste("v", i, sep = "."), lapply(lapply(Vlist,deriv, Istates[i]), eval, paras))
for(j in 1:length(Istates)){
k=k+1
fl[[k]]=attr(eval(as.name(paste("f", i, sep=".")))[[j]], "gradient")[1,]
vl[[k]]=attr(eval(as.name(paste("v", i, sep=".")))[[j]], "gradient")[1,]
}
}

f=matrix(as.numeric(as.matrix(fl)[,1]), ncol=length(Istates))
v=matrix(as.numeric(as.matrix(vl)[,1]), ncol=length(Istates))
R0=max(eigen(f%*%solve(v))$values)
return(R0)
}
```


Step 1: All states
```{r}
istates=c("I1", "I2")
```
Step 2: All new infections: 
```{r}
flist=c(dI1dt=quote(beta11 * S1 *(1-p1) * I1 / N1 + beta21 * S1 *(1-p1)* I2 / N1), dI2dt=quote(beta12 * S2*(1-p2) * I1 / N2 + beta22 * S2*(1-p2)* I2 / N2))
```
Step 3-5
```{r}
#All losses 
Vm1=quote(mu1 * I1 + gamma1 * I1)
Vm2=quote(mu2 * I2 + gamma2 * I2)

#All gained transfers$
Vp1=0
Vp2=0
#Subtract Vp from Vm
V1=substitute(a-b, list(a=Vm1, b=Vp1))
V2=substitute(a-b, list(a=Vm2, b=Vp2))
#Make Vlist
vlist = c(V1,V2)
```
Define list of parameter vectors:
```{r}
parms  = list(mu = c(0.2, 0.1), 
  N = c(1,1), 
  beta =  matrix(c(3, 2, 1, 1), ncol=2, byrow=TRUE), 
  gamma = c(1,1),
  p=c(0, 0))
#Extract to work with nextgenR0:
para = list(mu1 = parms$mu[1],
    mu2 = parms$mu[2],
    beta11 = parms$beta[1,1],
    beta12 = parms$beta[1,2],
    beta21 = parms$beta[2,1],
    beta22 = parms$beta[2,2],
    gamma1 = parms$gamma[1],
    gamma2 = parms$gamma[2],
    p1 = parms$p[1],
    p2 = parms$p[2],
    N1 = parms$N[1],
    N2 = parms$N[2])
```
Specify disease-free equilibrium
```{r}
df = list(S1 = 1, S2 = 1, I1 = 0, I2 = 0)
```
Invoke R0 calculator:
```{r}
nextgenR0(Istates=istates, Flist=flist, Vlist=vlist, params=para, dfe=df)
```


This next section generates an $R_E$ surface assuming that only the core species is vaccinated. For simplicity back-spill rates ($\beta_{2,1}$) are assumed identical to the intraspecific rate ({\beta_{2,2}}$) in the peripheral host.

```{r}
parms  = list(mu = c(0.2, 0.1), 
  N = c(1,1), 
  beta =  matrix(c(3, 2, 1, 1), ncol=2, byrow=TRUE), 
  gamma = c(1,1),
  p=c(0, 0))
#Extract to work with nextgenR0:
para = list(mu1 = parms$mu[1],
    mu2 = parms$mu[2],
    beta11 = parms$beta[1,1],
    beta12 = parms$beta[1,2],
    beta21 = parms$beta[2,1],
    beta22 = parms$beta[2,2],
    gamma1 = parms$gamma[1],
    gamma2 = parms$gamma[2],
    p1 = parms$p[1],
    p2 = parms$p[2],
    N1 = parms$N[1],
    N2 = parms$N[2])

beta2=seq(0,1, by=.02)
p=seq(0,1, by=0.02)
RE=matrix(NA, ncol=length(beta2), nrow=length(p))
for(i in 1:length(beta2)){
  for(j in 1:length(p)){
  para$p1=p[j]
  para$beta21=para$beta21=beta2[i]
  RE[i,j]=nextgenR0(Istates=istates, Flist=flist, Vlist=vlist, params=para, dfe=df)
}
}
image(beta2, p, RE)
contour(beta2, p, RE, levels=1:4, add=TRUE)
```