---
title: "jacobian"
author: "Ottar N. Bjornstad and Aaron King"
date: "11/02/2021"
output: html_document
---


Version 0.1.3 January 31, 2020
https://github.com/objornstad

Version 0.1.4 November 02, 2021


This Rmarkdown of a general purpose jacobian calculator was written by Ottar N. Bjørnstad and is released with a CC-BY-NC lisence for anyone to improve and re-share (acknowledging origin). Please email me a copy of update (onb1 at psu dot edu). The app was developed as part of the epimdr-project (https://cran.r-project.org/package=epimdr; Bjørnstad 2018).

**MOTIVATION** For stability analysis, ressonant periodicity, transfer-functions, etc, etc we usually need the Jacobian matrix evaluated with parameters and at some point in the phase plane.

A general purpose function is (with Aaron King's one-upmanship at the bottom):

```{r}
jacobian=function(states, elist, params, pts){
paras = as.list(c(pts, params)) 

k=0
jl=list(NULL)
for(i in 1:length(states)){
assign(paste("jj", i, sep = "."), lapply(lapply(elist, deriv, states[i]), eval, paras))
for(j in 1:length(states)){
k=k+1
jl[[k]]=attr(eval(as.name(paste("jj", i, sep=".")))[[j]], "gradient")[1,]
}
}

J=matrix(as.numeric(as.matrix(jl)[,1]), ncol=length(states))
return(J)
}
```


**states** is a vector naming all *state variables*, 

**elist** is a list that contains equations (**quotes**) for all states,

**params** is a *labeled vector* of parameters,

**pts** is a a *labeled vector* of the point in the phase plane to evaluate the Jacobian (often the endemic or disease-free equilibrium if working in mathematical epidemiology; or some other equilibrium if working in ecology). 

____
EXAMPLE 1 SIR (Bjornstad 2018: page 21)

Lets consider the SIR example. The basic equations for the flow of hosts between **S**usceptible, **I**nfectious and **R**ecovered
 compartments are:

\begin{align}
    \frac{dS}{dt} =& \underbrace{\mu N}_{\mbox{birth}} - \underbrace{\beta I \frac{S}{N}}_{\mbox{infection}} - \underbrace{\mu S}_{\mbox{death}} \label{eq:sirs}\\
     \frac{dI}{dt} =& \underbrace{\beta I \frac{S}{N}}_{\mbox{infection}} - \underbrace{\gamma I}_{\mbox{recovery}} - \underbrace{\mu I}_{\mbox{death}}  \label{eq:siri}\\
     \frac{dR}{dt} =& \underbrace{\gamma I}_{\mbox{recovery}} - \underbrace{\mu R}_{\mbox{death}} \label{eq:sirr}
\end{align}

Infected individuals infectious for an average time of $1/(\gamma+\mu)$ time units. The transmission rate is $\beta$. Because **R** is absobing and does not affect dynamics when we work on compartementalfractions ($N = 1$) we omit this equation.

Step 1: classes are S and I

```{r}
states=c("S", "I")
```

Step 2: Equations are:
```{r}
elist=c(dSdt = quote(mu * (N  - S)  - beta * S * I / N),
dIdt = quote(beta * S * I / N - (mu + gamma) * I))
```

Step 3: Some arbitrary parameters are:

```{r}
parms  = c(mu = 1/(50*52), N = 1, beta =  2, 
      gamma = 1/2)
```

Step 4: for this model the endemic equilibrium is $\{S^∗=\beta/(\gamma+\mu),I^∗=\gamma∗(\beta/(\gamma+\mu)−1)/\beta\}$
 and the disease-free equilibrium is $\{S^∗=1,I^∗=0\}$
 
```{r}
eeq=with(as.list(parms), c(I=(gamma+mu)/beta, S=mu*(beta/(gamma+mu)-1)/beta))
deq = list(S = 1, I = 0, R = 0)
```
Invoke Jacobian calculator:

```{r}
JJ=jacobian(states=states, elist=elist, params=parms, pts=eeq)
JJ
```
Eigen values are:
```{r}
eigen(JJ)$value
```
A pair of comnplex conjugates. So the endemic equilibrium is a stable focus. The ressonant periodicity is:
```{r}
2*pi/Im(eigen(JJ)$value[1])
```
We next look at the disease-free equilibrium:

```{r}
deq = list(S = 1, I = 0, R = 0)
JJ=jacobian(states=states, elist=elist, params=parms, pts=deq)
#Eigen values are:
eigen(JJ)$value
```
The dominant EV is real, so dfe is unstable (R0>1).
_______
EXAMPLE 2: SEIR

the SEIR model of the  flow of hosts between **S**usceptible, **E**xposed (but not yet
infectious), **I**nfectious and **R**ecovered compartments in a randomly mixing population:

\begin{align}
  \frac{dS}{dt} &=& \underbrace{\mu (N[1-p])}_{\mbox{recruitment}} -\underbrace{\beta  I \frac{S}{N}}_{\mbox{infected}} -\underbrace{\mu S}_{\mbox{dead}}
  \label{eq:seirs}\\
 \frac{dE}{dt} &=& \underbrace{\beta I \frac{S}{N}}_{\mbox{infected}} - \underbrace{\sigma E}_{\mbox{infectious}} - \underbrace{\mu I}_{\mbox{dead}}  \label{eq:seire}\\
\frac{dI}{dt} &=& \underbrace{\sigma E}_{\mbox{infectious}} - \underbrace{\gamma I}_{\mbox{recovered}} - \underbrace{(\mu +\alpha) I}_{\mbox{dead}}
  \label{eq:seiri}\\
  \frac{dR}{dt} &=& \underbrace{\gamma I}_{\mbox{recovered}} - \underbrace{\mu R}_{\mbox{dead}} + \underbrace{\mu N p}_{\mbox{vaccinated}}
  \label{eq:seirr}
\end{align}

where, susceptibles are either vaccinated at birth (fraction $p$), or infected at a rate $\beta I / N$.
Infected individuals will remain in the latent class for an average period of $1/(\sigma +\mu)$ and
 subsequently (if they escape natural mortality at a rate $\mu$) enter the infectious class for 
 an average time of $1/(\gamma+\mu+\alpha)$; $\alpha$ is the *rate* of disease induced mortality (*not* case fatality rate). By the rules of competing rates , the case fatality rate is $\alpha/(\gamma+\mu+\alpha)$ because during the time an individual is expected to remain in the infectious class the disease is killing them at a rate $\alpha$. By a similar logic, the probability of recovering with immunity (for life in the case of the SEIR model) is $\gamma/(\gamma+\mu+\alpha)$. Putting all these pieces together, and assuming no vaccination, the expected number of secondary cases in a completely susceptible population is thus: probability of making it through latent stage without dying * expected infectious period * transmission rate while infectious. Thus, $R_0 =  \frac{\sigma}{\sigma +\mu} \frac{1}{\gamma+\mu+\alpha} \frac{\beta N}{N} =  \frac{\sigma}{\sigma +\mu} \frac{\beta}{\gamma+\mu+\alpha}$.

```{r}
states2=c("S", "E", "I", "R")
```


```{r}
elist2=c(dS = quote(mu * (N  - S)  - beta * S * I / N),
  dE = quote(beta * S * I / N - (mu + sigma) * E),
  dI = quote(sigma * E - (mu + gamma+alpha) * I),
 dR = quote(gamma * I - mu * R))
```



```{r}
paras2  = c(mu = 1/50, N = 1, beta =  1000, 
     sigma = 365/8, gamma = 365/5, alpha=0)
```

```{r}
deq2=list(S = 1, E = 0, I = 0, R = 0)
```


```{r}
JJ=jacobian(states=states2, elist=elist2, params=paras2, pts=deq2)
eigen(JJ)$value
```
Largest eigenvalue is positive and strictly real so the disease-free equilibrium is an unstable node.


______
EXAMPLE 3: The Rosenzweig-MacArthur predator-prey model. 

The basic equations for the  consumer-resource interaction between prey (N)  and predators (P) are:
\begin{align}
    \frac{dN}{dt} =& \underbrace{r N (\frac{K-N}{K})}_{\mbox{N growth}} - \underbrace{\frac{a N P}{c + N}}_{\mbox{predation}}\\
     \frac{dP}{dt} =& \underbrace{\frac{b N P}{c + N}}_{\mbox{P growth}} - \underbrace{g P}_{\mbox{P death}}
\end{align}

Prey ($N$) are assumed to grow acording to the logistic model with a maximum growth rate, $r$ and carrying-capacity, $K$. Predators ($P$)are feeding according to a Type-II functional respose with a maximum search efficiency, $a$ and half-saturation constant $c$. Predators have a conversion efficiency of $b/a$ and a life-expectancy of $1/g$.

The isoclines (sometimes called the nullclines) of this system are given by the solution to the 
equations $\frac{dN}{dt} = 0$ and $\frac{dP}{dt} = 0$ and partitions the phase plane into regions 
were $N$ and $P$ are increasing and decreasing. The $N$-isocline is $P^* = (r-rN/K)(c+N)/a$
and the P-isocline is $N^* = gc/(b-g)$. The equilibrium is
\begin{align}
N^* =& gc/(b-g) \\
P^* =& (r-rN^*/K)(c+N^*)/a
\end{align}

```{r}
states3=c("N", "P")
```


```{r}
elist3=c(dN = quote(r *N *((K-N)/K) - a *N *P/(c + N)),
     dP =quote(b*N*P/(c + N) - g*P))
```

```{r}
paras3  = c(r=0.1, K=90, a=0.2, c=20, b=0.1, g=0.05)
```

```{r}
eq3=with(as.list(paras3), c(N=g*c/(b-g), P=(r-r*g*c/(b-g)/K)*(c+g*c/(b-g))/a))
```

```{r}
JJ=jacobian(states=states3, elist=elist3, params=paras3, pts=eq3)
eigen(JJ)$value
```
Equilibrium is an unstable focus with damping period of:
```{r}
2*pi/Im(eigen(JJ)$value[1])
```

Aarons one-upmanship:
"I can see how useful it will be.  This kind of programming in R seems unavoidably tortuous, but it is amazing how far one can push things in the general lisp-like structure of R.

I can't read code without tinkering with it, so I played around a bit with your Jacobian calculator.  I thought it might be interesting to have one that returns a function for the Jacobian that can be evaluated at multiple points.  Then I thought I'd see how far I could go in terms of making the calling syntax simpler.  What do you think of the attached?

This will only run on the latest R (4.1) because of the native pipes it uses in places.  It would be trivial to remove those, but as you know I am addicted to pipes.  (I got absurdly excited about 4.1)."




```{r}
stopifnot(getRversion()>="4.1")

Jacobian <- function (.vars, ...) {
  vf <- substitute(list(...))[-1L]
  vars <- sapply(substitute(.vars),deparse)
  if (length(vars)>1) vars <- vars[-1L]
  sapply(
    vars,
    \(var) sapply(vf,D,name=var)
  ) -> jac
  dd <- c(length(vf),length(vars))
  dim(jac) <- NULL
  dn <- list(
    ifelse(
      nzchar(names(vf)),
      names(vf),
      sapply(vf,deparse)
    ),
    vars
  )
  fun <- function (...) {
    e <- c(as.list(sys.frame(sys.nframe())),...)
    J <- vapply(jac,eval,numeric(1L),envir=e)
    dim(J) <- dd
    dimnames(J) <- dn
    J
  }
  formals(fun) <- eval(
    parse(text=paste0("alist(",paste0(c(vars,"..."),"=",collapse=","),")"))
  )
  fun
}
```

Some examples:
```{r}
Jacobian(c(x,y),sin(x),cos(x),atan(y/x),tan(x+y)) -> f
f
f()
f(3,2)
f(y=2,x=3)
```

```{r}
Jacobian(x,sin(x),cos(x),atan(y/x),tan(x+y),factorial(x)) -> f
f
f(x=3)
f(x=3,y=2)
```

```{r}
Jacobian(c(x,y),A=sin(x),B=cos(x),atan(y/x),D=tan(x+y),factorial(x)) -> f
f(y=3,x=2)
```

```{r}
c(S,I,R) |>
  Jacobian(
    dSdt=mu*(S+I+R)-beta*S*I/(S+I+R)-mu*S,
    dIdt=beta*S*I/(S+I+R)-gamma*I-(mu+alpha)*I,
    dRdt=gamma*I-mu*R
  ) -> f
```

f
f(S=0.99,I=0.01,R=0,mu=0.02,beta=400,gamma=26,alpha=1)
with(
  list(gamma=26,beta=400,alpha=1,mu=0.02),
  f(
    S=(gamma+mu)/beta,
    I=(1-(gamma+mu)/beta)*mu/(gamma+mu),
    R=(1-(gamma+mu)/beta)*gamma/(gamma+mu),
    mu=0.02,beta=400,gamma=26,alpha=1
  )
)|>
  eigen()

```
