---
title: "Lotka-Volterra predation model"
author: Ottar N. Bjørnstad
output: html_document
runtime: shiny
---

Version 0.5.3 August 26, 2019 
https://github.com/objornstad/

This Rmarkdown of the Lotka-Volterra predation model was written by Ottar N. Bjørnstad and is released with a CC-BY-NC license for anyone to improve and re-share (acknowledging origin). Please email me a copy if you make an update (onb1 at psu dot edu).

The app requires the shiny, and deSolve packages to be installed to run. 
```{r, echo=FALSE}
require(shiny)
require(deSolve)
#The flowField function was written by Michael J Grayling (https://github.com/mjg211) as part of 
#recently archived phaseR-package and is reproduced here under the original GPL-3 license
flowField=function (deriv, xlim, ylim, parameters = NULL, system = "two.dim", 
   points = 21, col = "gray", arrow.type = "equal", arrow.head = 0.05, 
    frac = 1, add = TRUE, xlab = "x", ylab = "y", state.names = c("x", 
        "y"), ...) 
{
    if ((!is.vector(xlim)) | (length(xlim) != 2)) {
        stop("xlim is not a vector of length 2 as required")
    }
    if (xlim[2] <= xlim[1]) {
        stop("xlim[2] is less than or equal to xlim[1]")
    }
    if ((!is.vector(ylim)) | (length(ylim) != 2)) {
        stop("ylim is not a vector of length 2 as required")
    }
    if (ylim[2] <= ylim[1]) {
        stop("ylim[2] is less than or equal to ylim[1]")
    }
    if (points <= 0) {
        stop("points is less than or equal to zero")
    }
    if (!(system %in% c("one.dim", "two.dim"))) {
        stop("system must either be set to \"one.dim\" or \"two.dim\"")
    }
    if (is.vector(col) == FALSE) {
        stop("col is not a vector as required")
    }
    if (length(col) > 1) {
        col <- col[1]
        print("Note: col has been reset as required")
    }
    if (!(arrow.type %in% c("proportional", "equal"))) {
        stop("arrow.type must either be set to \"proportional\" or \"equal\"")
    }
    if (arrow.head <= 0) {
        stop("arrow.head is less than or equal to zero")
    }
    if (frac <= 0) {
        stop("frac is less than or equal to zero")
    }
    if (!is.logical(add)) {
        stop("add must be logical")
    }
    x <- seq(from = xlim[1], to = xlim[2], length = points)
    y <- seq(from = ylim[1], to = ylim[2], length = points)
    dx <- matrix(0, ncol = points, nrow = points)
    dy <- matrix(0, ncol = points, nrow = points)
    xmax.length <- x[2] - x[1]
    ymax.length <- y[2] - y[1]
    if (add == FALSE) {
        plot(1, xlim = c(xlim[1] - xmax.length, xlim[2] + xmax.length), 
            ylim = c(ylim[1] - ymax.length, ylim[2] + ymax.length), 
            type = "n", xlab = xlab, ylab = ylab, ...)
    }
    if (system == "one.dim") {
        for (i in 1:points) {
            dy[1, i] <- deriv(0, setNames(c(y[i]), state.names[1]), 
                parameters)[[1]]
        }
        for (i in 2:points) {
            dy[i, ] <- dy[1, ]
        }
        abs.dy <- abs(dy)
        abs.dy.non <- abs.dy[which(abs.dy != 0)]
        max.abs.dy <- max(abs(dy))
        coefficient <- frac * min(xmax.length, ymax.length)/(2 * 
            sqrt(2) * max(sqrt(2 * abs.dy.non/(abs.dy.non + (1/abs.dy.non))), 
            sqrt(2 * (1/abs.dy.non)/(abs.dy.non + (1/abs.dy.non)))))
        for (i in 1:points) {
            for (j in 1:points) {
                if (dy[i, j] != 0) {
                  factor <- sqrt(2/(abs.dy[i, j] + (1/abs.dy[i, 
                    j])))
                  y.shift <- coefficient * factor * sqrt(abs.dy[i, 
                    j])
                  x.shift <- coefficient * factor/sqrt(abs.dy[i, 
                    j])
                  if (dy[i, j] < 0) {
                    y.shift <- -y.shift
                  }
                }
                if (dy[i, j] == 0) {
                  y.shift <- 0
                  x.shift <- coefficient * sqrt(2)
                }
                if (arrow.type == "proportional") {
                  if (dy[i, j] != 0) {
                    prop <- abs.dy[i, j]/max.abs.dy
                    y.shift <- y.shift * prop
                    x.shift <- x.shift * prop
                  }
                  if (dy[i, j] == 0) {
                    x.shift <- y.shift * mean(abs.dy)/max.abs.dy
                  }
                }
                arrows(x[i] - x.shift, y[j] - y.shift, x[i] + 
                  x.shift, y[j] + y.shift, length = arrow.head, 
                  col = col, ...)
            }
        }
        return(list(add = add, arrow.head = arrow.head, arrow.type = arrow.type, 
            col = col, deriv = deriv, dy = dy, frac = frac, parameters = parameters, 
            points = points, system = system, x = x, xlab = xlab, 
            xlim = xlim, y = y, ylab = ylab, ylim = ylim))
    }
    else {
        for (i in 1:length(x)) {
            for (j in 1:length(y)) {
                df <- deriv(0, setNames(c(x[i], y[j]), state.names), 
                  parameters)
                dx[i, j] <- df[[1]][1]
                dy[i, j] <- df[[1]][2]
            }
        }
        abs.dx <- abs(dx)
        abs.dy <- abs(dy)
        abs.dx.non <- abs.dx[which((abs.dx != 0) & (abs.dy != 
            0))]
        abs.dy.non <- abs.dy[which((abs.dx != 0) & (abs.dy != 
            0))]
        max.length <- max(sqrt(dx^2 + dy^2))
        coefficient <- frac * min(xmax.length, ymax.length)/(2 * 
            sqrt(2) * max(sqrt(2 * (abs.dy.non/abs.dx.non)/((abs.dy.non/abs.dx.non) + 
            (abs.dx.non/abs.dy.non))), sqrt(2 * (abs.dx.non/abs.dy.non)/((abs.dy.non/abs.dx.non) + 
            (abs.dx.non/abs.dy.non)))))
        for (i in 1:points) {
            for (j in 1:points) {
                if ((dx[i, j] != 0) | (dy[i, j] != 0)) {
                  if ((dx[i, j] != 0) & (dy[i, j] != 0)) {
                    factor <- sqrt(2/((abs.dy[i, j]/abs.dx[i, 
                      j]) + (abs.dx[i, j]/abs.dy[i, j])))
                    y.shift <- coefficient * factor * sqrt(abs.dy[i, 
                      j]/abs.dx[i, j])
                    x.shift <- coefficient * factor/sqrt(abs.dy[i, 
                      j]/abs.dx[i, j])
                    if (dy[i, j] < 0) {
                      y.shift <- -abs(y.shift)
                    }
                    if (dx[i, j] < 0) {
                      x.shift <- -abs(x.shift)
                    }
                  }
                  if ((dx[i, j] == 0) & (dy[i, j] != 0)) {
                    y.shift <- coefficient * sqrt(2)
                    x.shift <- 0
                    if (dy[i, j] < 0) {
                      y.shift <- -abs(y.shift)
                    }
                  }
                  if ((dx[i, j] != 0) & (dy[i, j] == 0)) {
                    y.shift <- 0
                    x.shift <- coefficient * sqrt(2)
                    if (dx[i, j] < 0) {
                      x.shift <- -abs(x.shift)
                    }
                  }
                  if (arrow.type == "proportional") {
                    prop <- sqrt((abs.dx[i, j]^2 + abs.dy[i, 
                      j]^2))/max.length
                    y.shift <- y.shift * prop
                    x.shift <- x.shift * prop
                  }
                  arrows(x[i] - x.shift, y[j] - y.shift, x[i] + 
                    x.shift, y[j] + y.shift, length = arrow.head, 
                    col = col, ...)
                }
            }
        }
    }
    return(list(add = add, arrow.head = arrow.head, arrow.type = arrow.type, 
        col = col, deriv = deriv, dx = dx, dy = dy, frac = frac, 
        parameters = parameters, points = points, system = system, 
        x = x, xlab = xlab, xlim = xlim, y = y, ylab = ylab, 
        ylim = ylim))
}

```

The basic equations for the consumer-resource interaction between prey (N)  and predators (P) are:
\begin{align}
    \frac{dN}{dt} =& \underbrace{r N)}_{\mbox{N growth}} - \underbrace{a N P}_{\mbox{predation}}\\
     \frac{dP}{dt} =& \underbrace{b a N P}_{\mbox{P growth}} - \underbrace{m P}_{\mbox{P death}}
\end{align}

Prey are assumed to grow exponentially at a rate, $r$. Predators are feeding at a per-capita rate $a$. Predators have a conversion efficiency of $b$ and a life-expectancy of $1/m$.

The isoclines (sometimes called the nullclines) of this system are given by the solution to the 
equations $dN/dt=0$ and $dP/dt=0$ and partitions the phase plane into regions 
were $N$ and $P$ are increasing and decreasing. 

The $N$-isocline is $$P^* = r/a$$
and the $P$-isocline is $$N^* = g/b$$ 

The  shiny app:

```{r, echo=FALSE}
# This creates the User Interface (UI)
ui = fluidPage(
  tags$head(tags$style(
    HTML("
         #sidebar1 {
            background-color: #ECECEC;
        }
    
    #sidebar2 {
      background-color: #ECECEC
    }")
  )),
titlePanel(""),
fluidRow(
column(4, id = "sidebar2",
fluidRow(column(5, id = "sidebar1",
sliderInput("r", "r:", 0.6,
              min = 0, max = 1, step=0.01),
sliderInput("a", "a:", 0.1,
              min = 0, max = 1, step=0.01),
numericInput("N0", "initial N:", 10,
              min = 0, max = 100)),
column(5, offset = 1, id = "sidebar1",
sliderInput("b", "b:", 0.1,
              min = 0, max = 1, step=0.01),
sliderInput("m", "m:", 0.2,
               min = 0, max = 1, step=0.01),
numericInput("P0", "initial P:", 10,
              min = 0, max = 100)),
column(1)),
fluidRow(
column(6, offset = 3, id = "sidebar1",
numericInput("Tmax", "Tmax:", 100,
               min = 0, max = 200)),
column(3))
),
mainPanel(tabsetPanel(
  tabPanel("Simulation", plotOutput("plot1", height = 500)),
     tabPanel("Details", 
     withMathJax(
                helpText("MODEL:"),
          helpText("Prey $$\\frac{dN}{dt} = r N - a N P$$"),
          helpText("Predator $$\\frac{dP}{dt} = b a N P - m P$$"),
          helpText("N-isocline $$P^* = r/a$$"),
          helpText("P-isocline $$N^* = g/b$$"),
          helpText("Equilibrium $$N^* = g/b, P^* = r/a$$")))
           )
)
)
)

# This creates the "behind the scenes" code (Server)
server = function(input, output) {
LV=function(t, y, parameters){
  N=y[1]
  P=y[2]

  with(as.list(parameters),{
  dN = r*N-a*N*P
  dP =b*a*N*P-m*P
  res=c(dN, dP)
  list(res)
})
}

output$plot1 <- renderPlot({
Nstar=input$m/(input$b*input$a)
Pstar=input$r/input$a

times  = seq(0, input$Tmax, by=0.1)
parms=c(r=input$r,a=input$a,b=input$b,m=input$m)
xstart = c(N=input$N0, P=input$P0)
 
out=ode(y=xstart,
  times=times,
  func=LV,
  parms=parms)

  out=as.data.frame(out)

 
par(mfrow=c(1,2))  #This puts two plots side by side each other
plot(times, out$N, ylab="abundance", xlab="time", type="l", ylim=range(out[,1:2]))
lines(times, out$P, col="red")
  legend("right",
        legend=c("N", "P"),
        lty=c(1,1),
         col=c("black", "red"))

plot(out$N, out$P, ylab="Predator (P)", xlab="Prey (N)", type="l", xlim=range(out[,2]), ylim=range(out[,3]))
abline(h=Pstar, col = "black")
abline(v=Nstar, col = "red")
fld=flowField(LV, xlim=range(out[,2]), ylim=range(out[,3]), 
parameters=parms, system="two.dim", add=TRUE)
points(0,0,pch = 1)
points(Nstar,Pstar, pch = 19)
   })
  }

shinyApp(ui, server, options = list(height = 550))
```

Reference: