---
title: "Lotka-Volterra predation model"
author: Ottar N. Bjørnstad
output: html_document
runtime: shiny
---

Version 0.5 July 1, 2019 
https://github.com/objornstad/

This Rmarkdown of the Lotka-Volterra predation model was written by Ottar N. Bjørnstad and is released with a CC-BY-NC license for anyone to improve and re-share (acknowledging origin). Please email me a copy if you make an update (onb1 at psu dot edu).

The app requires the shiny, deSolve and phaseR packages to be installed to run. 
```{r}
require(shiny)
require(deSolve)
require(phaseR)
```

The basic equations for the consumer-resource interaction between prey (N)  and predators (P) are:
\begin{align}
    \frac{dN}{dt} =& \underbrace{r N)}_{\mbox{N growth}} - \underbrace{a N P}_{\mbox{predation}}\\
     \frac{dP}{dt} =& \underbrace{b a N P}_{\mbox{P growth}} - \underbrace{m P}_{\mbox{P death}}
\end{align}

Prey are assumed to grow exponentially at a rate, $r$. Predators are feeding at a per-capita rate $a$. Predators have a conversion efficiency of $b$ and a life-expectancy of $1/m$.

The isoclines (sometimes called the nullclines) of this system are given by the solution to the 
equations $dN/dt=0$ and $dP/dt=0$ and partitions the phase plane into regions 
were $N$ and $P$ are increasing and decreasing. 

The $N$-isocline is $$P^* = r/a$$
and the $P$-isocline is $$N^* = g/b$$ 

The  shiny app:

```{r, echo=FALSE}
# This creates the User Interface (UI)
ui = fluidPage(
  tags$head(tags$style(
    HTML("
         #sidebar1 {
            background-color: #ECECEC;
        }
    
    #sidebar2 {
      background-color: #ECECEC
    }")
  )),
titlePanel(""),
fluidRow(
column(4, id = "sidebar2",
fluidRow(column(5, id = "sidebar1",
sliderInput("r", "r:", 0.6,
              min = 0, max = 1, step=0.01),
sliderInput("a", "a:", 0.1,
              min = 0, max = 1, step=0.01),
numericInput("N0", "initial N:", 10,
              min = 0, max = 100)),
column(5, offset = 1, id = "sidebar1",
sliderInput("b", "b:", 0.1,
              min = 0, max = 1, step=0.01),
sliderInput("m", "m:", 0.2,
               min = 0, max = 1, step=0.01),
numericInput("P0", "initial P:", 10,
              min = 0, max = 100)),
column(1)),
fluidRow(
column(6, offset = 3, id = "sidebar1",
numericInput("Tmax", "Tmax:", 100,
               min = 0, max = 200)),
column(3))
),
mainPanel(tabsetPanel(
  tabPanel("Simulation", plotOutput("plot1", height = 500)),
     tabPanel("Details", 
     withMathJax(
                helpText("MODEL:"),
          helpText("Prey $$\\frac{dN}{dt} = r N - a N P$$"),
          helpText("Predator $$\\frac{dP}{dt} = b a N P - m P$$"),
          helpText("N-isocline $$P^* = r/a$$"),
          helpText("P-isocline $$N^* = g/b$$"),
          helpText("Equilibrium $$N^* = g/b, P^* = r/a$$")))
           )
)
)
)

# This creates the "behind the scenes" code (Server)
server = function(input, output) {
LV=function(t, y, parameters){
  N=y[1]
  P=y[2]

  with(as.list(parameters),{
  dN = r*N-a*N*P
  dP =b*a*N*P-m*P
  res=c(dN, dP)
  list(res)
})
}

output$plot1 <- renderPlot({
Nstar=input$m/(input$b*input$a)
Pstar=input$r/input$a

times  = seq(0, input$Tmax, by=0.1)
parms=c(r=input$r,a=input$a,b=input$b,m=input$m)
xstart = c(N=input$N0, P=input$P0)
 
out=ode(y=xstart,
  times=times,
  func=LV,
  parms=parms)

  out=as.data.frame(out)

 
par(mfrow=c(1,2))  #This puts two plots side by side each other
plot(times, out$N, ylab="abundance", xlab="time", type="l", ylim=range(out[,1:2]))
lines(times, out$P, col="red")
  legend("right",
        legend=c("N", "P"),
        lty=c(1,1),
         col=c("black", "red"))

plot(out$N, out$P, ylab="Predator (P)", xlab="Prey (N)", type="l", xlim=range(out[,2]), ylim=range(out[,3]))
abline(h=Pstar, col = "black")
abline(v=Nstar, col = "red")
fld=flowField(LV, xlim=range(out[,2]), ylim=range(out[,3]), 
parameters=parms, system="two.dim", add=TRUE)
points(0,0,pch = 1)
points(Nstar,Pstar, pch = 19)
   })
  }

shinyApp(ui, server)
```

Reference: