---
title: "Lotka-Volterra competion model"
author: Ottar N. Bjørnstad
output: html_document
runtime: shiny
---

Version 0.5.3 August 26, 2019 
https://github.com/objornstad/

This Rmarkdown of the Lotka-Volterra competition model was written by Ottar N. Bjørnstad and is released with a CC-BY-NC license for anyone to improve and re-share (acknowledging origin). Please email me a copy of update (onb1 at psu dot edu).

The app requires the shiny and deSolve packages to be installed to run. 

```{r, echo=FALSE}
require(shiny)
require(deSolve)
#The flowField function was written by Michael J Grayling (https://github.com/mjg211) as part of 
#recently archived phaseR-package and is reproduced here under the original GPL-3 license
flowField=function (deriv, xlim, ylim, parameters = NULL, system = "two.dim", 
   points = 21, col = "gray", arrow.type = "equal", arrow.head = 0.05, 
    frac = 1, add = TRUE, xlab = "x", ylab = "y", state.names = c("x", 
        "y"), ...) 
{
    if ((!is.vector(xlim)) | (length(xlim) != 2)) {
        stop("xlim is not a vector of length 2 as required")
    }
    if (xlim[2] <= xlim[1]) {
        stop("xlim[2] is less than or equal to xlim[1]")
    }
    if ((!is.vector(ylim)) | (length(ylim) != 2)) {
        stop("ylim is not a vector of length 2 as required")
    }
    if (ylim[2] <= ylim[1]) {
        stop("ylim[2] is less than or equal to ylim[1]")
    }
    if (points <= 0) {
        stop("points is less than or equal to zero")
    }
    if (!(system %in% c("one.dim", "two.dim"))) {
        stop("system must either be set to \"one.dim\" or \"two.dim\"")
    }
    if (is.vector(col) == FALSE) {
        stop("col is not a vector as required")
    }
    if (length(col) > 1) {
        col <- col[1]
        print("Note: col has been reset as required")
    }
    if (!(arrow.type %in% c("proportional", "equal"))) {
        stop("arrow.type must either be set to \"proportional\" or \"equal\"")
    }
    if (arrow.head <= 0) {
        stop("arrow.head is less than or equal to zero")
    }
    if (frac <= 0) {
        stop("frac is less than or equal to zero")
    }
    if (!is.logical(add)) {
        stop("add must be logical")
    }
    x <- seq(from = xlim[1], to = xlim[2], length = points)
    y <- seq(from = ylim[1], to = ylim[2], length = points)
    dx <- matrix(0, ncol = points, nrow = points)
    dy <- matrix(0, ncol = points, nrow = points)
    xmax.length <- x[2] - x[1]
    ymax.length <- y[2] - y[1]
    if (add == FALSE) {
        plot(1, xlim = c(xlim[1] - xmax.length, xlim[2] + xmax.length), 
            ylim = c(ylim[1] - ymax.length, ylim[2] + ymax.length), 
            type = "n", xlab = xlab, ylab = ylab, ...)
    }
    if (system == "one.dim") {
        for (i in 1:points) {
            dy[1, i] <- deriv(0, setNames(c(y[i]), state.names[1]), 
                parameters)[[1]]
        }
        for (i in 2:points) {
            dy[i, ] <- dy[1, ]
        }
        abs.dy <- abs(dy)
        abs.dy.non <- abs.dy[which(abs.dy != 0)]
        max.abs.dy <- max(abs(dy))
        coefficient <- frac * min(xmax.length, ymax.length)/(2 * 
            sqrt(2) * max(sqrt(2 * abs.dy.non/(abs.dy.non + (1/abs.dy.non))), 
            sqrt(2 * (1/abs.dy.non)/(abs.dy.non + (1/abs.dy.non)))))
        for (i in 1:points) {
            for (j in 1:points) {
                if (dy[i, j] != 0) {
                  factor <- sqrt(2/(abs.dy[i, j] + (1/abs.dy[i, 
                    j])))
                  y.shift <- coefficient * factor * sqrt(abs.dy[i, 
                    j])
                  x.shift <- coefficient * factor/sqrt(abs.dy[i, 
                    j])
                  if (dy[i, j] < 0) {
                    y.shift <- -y.shift
                  }
                }
                if (dy[i, j] == 0) {
                  y.shift <- 0
                  x.shift <- coefficient * sqrt(2)
                }
                if (arrow.type == "proportional") {
                  if (dy[i, j] != 0) {
                    prop <- abs.dy[i, j]/max.abs.dy
                    y.shift <- y.shift * prop
                    x.shift <- x.shift * prop
                  }
                  if (dy[i, j] == 0) {
                    x.shift <- y.shift * mean(abs.dy)/max.abs.dy
                  }
                }
                arrows(x[i] - x.shift, y[j] - y.shift, x[i] + 
                  x.shift, y[j] + y.shift, length = arrow.head, 
                  col = col, ...)
            }
        }
        return(list(add = add, arrow.head = arrow.head, arrow.type = arrow.type, 
            col = col, deriv = deriv, dy = dy, frac = frac, parameters = parameters, 
            points = points, system = system, x = x, xlab = xlab, 
            xlim = xlim, y = y, ylab = ylab, ylim = ylim))
    }
    else {
        for (i in 1:length(x)) {
            for (j in 1:length(y)) {
                df <- deriv(0, setNames(c(x[i], y[j]), state.names), 
                  parameters)
                dx[i, j] <- df[[1]][1]
                dy[i, j] <- df[[1]][2]
            }
        }
        abs.dx <- abs(dx)
        abs.dy <- abs(dy)
        abs.dx.non <- abs.dx[which((abs.dx != 0) & (abs.dy != 
            0))]
        abs.dy.non <- abs.dy[which((abs.dx != 0) & (abs.dy != 
            0))]
        max.length <- max(sqrt(dx^2 + dy^2))
        coefficient <- frac * min(xmax.length, ymax.length)/(2 * 
            sqrt(2) * max(sqrt(2 * (abs.dy.non/abs.dx.non)/((abs.dy.non/abs.dx.non) + 
            (abs.dx.non/abs.dy.non))), sqrt(2 * (abs.dx.non/abs.dy.non)/((abs.dy.non/abs.dx.non) + 
            (abs.dx.non/abs.dy.non)))))
        for (i in 1:points) {
            for (j in 1:points) {
                if ((dx[i, j] != 0) | (dy[i, j] != 0)) {
                  if ((dx[i, j] != 0) & (dy[i, j] != 0)) {
                    factor <- sqrt(2/((abs.dy[i, j]/abs.dx[i, 
                      j]) + (abs.dx[i, j]/abs.dy[i, j])))
                    y.shift <- coefficient * factor * sqrt(abs.dy[i, 
                      j]/abs.dx[i, j])
                    x.shift <- coefficient * factor/sqrt(abs.dy[i, 
                      j]/abs.dx[i, j])
                    if (dy[i, j] < 0) {
                      y.shift <- -abs(y.shift)
                    }
                    if (dx[i, j] < 0) {
                      x.shift <- -abs(x.shift)
                    }
                  }
                  if ((dx[i, j] == 0) & (dy[i, j] != 0)) {
                    y.shift <- coefficient * sqrt(2)
                    x.shift <- 0
                    if (dy[i, j] < 0) {
                      y.shift <- -abs(y.shift)
                    }
                  }
                  if ((dx[i, j] != 0) & (dy[i, j] == 0)) {
                    y.shift <- 0
                    x.shift <- coefficient * sqrt(2)
                    if (dx[i, j] < 0) {
                      x.shift <- -abs(x.shift)
                    }
                  }
                  if (arrow.type == "proportional") {
                    prop <- sqrt((abs.dx[i, j]^2 + abs.dy[i, 
                      j]^2))/max.length
                    y.shift <- y.shift * prop
                    x.shift <- x.shift * prop
                  }
                  arrows(x[i] - x.shift, y[j] - y.shift, x[i] + 
                    x.shift, y[j] + y.shift, length = arrow.head, 
                    col = col, ...)
                }
            }
        }
    }
    return(list(add = add, arrow.head = arrow.head, arrow.type = arrow.type, 
        col = col, deriv = deriv, dx = dx, dy = dy, frac = frac, 
        parameters = parameters, points = points, system = system, 
        x = x, xlab = xlab, xlim = xlim, y = y, ylab = ylab, 
        ylim = ylim))
}

```

The basic equations for the competitive interaction between species 1 ($N_1$)  and species 2 ($N_2$) are:
\begin{align}
    \frac{dN_1}{dt} =& r_1 N_1 (\frac{K_1-N_1-\alpha_{12}N_2}{K_1})\\
    \frac{dN_2}{dt} =& r_1 N_2 (\frac{K_2-N_2-\alpha_{21}N_1}{K_2})\\
\end{align}

Both species are assumed to grow according to the logistic model with maximum growth rates $r_1$ and $r_2$, and carrying capacities, $K_1$ and $K_2$. Interspecific competiotion coefficents are $\alpha_{12}$ and $\alpha_{21}$.

The isoclines (sometimes called the nullclines) of this system -- given by the solution to the 
equations $dN_1/dt=0$ and $dN_2/dt=0$ -- partitions the phase plane into regions 
were $N_1$ and $N_2$ are increasing and decreasing. The $N_1$ isocline is
$$N_2 = \frac{K_1 - N_1}{\alpha_{12}}$$
The $N_2$ isocline is
$$N_2 = K_2 - \alpha_{21} N_1$$
Depending on parameters the equilibria are $$\{N_1^*=K_1,N_2^*=0\}$$, $$\{N_1^*=0, N_2^*=K_2\}$$ or 
$$\{N_1^* = \frac{K_1-\alpha_{12} K_2}{1-\alpha_{12} \alpha_{21}},
           N_2^* = \frac{K_2-\alpha_{21} K_1}{1-\alpha_{12} \alpha_{21}}\}$$

The  shiny app:

```{r, echo=FALSE}
ui = fluidPage(
# This creates the User Interface (UI)
  tags$head(tags$style(
    HTML('
         #sidebar1 {
            background-color: #ECECEC;
        }
    
    #sidebar2 {
      background-color: #ECECEC
    }')
  )),
titlePanel(""),
fluidRow(
column(4, id = "sidebar2",
fluidRow(column(5, id = "sidebar1",
sliderInput("r1", "r1:", 0.3,
               min = 0, max = 1, step=0.01),
sliderInput("a", "alpha12:", 0.3,
               min = 0, max = 1, step=0.01),
sliderInput("K1", "K1:", 150,
               min = 0, max = 300, step=1),
numericInput("N1", "initial N1:", 10,
               min = 0, max = 1000)),
column(5, offset = 1, id = "sidebar1",
sliderInput("r2", "r2:", 0.5,
               min = 0, max = 1, step=0.01),
sliderInput("b", "alpha21:", 0.45,
               min = 0, max = 1, step=0.01),
sliderInput("K2", "K2:", 100,
               min = 0, max = 300, step=1),
numericInput("N2", "initial N2:", 15,
               min = 0, max = 1000)),
column(1)),
fluidRow(
column(6, offset = 3, id = "sidebar1",
numericInput("Tmax", "Tmax:", 100,
               min = 0, max = 200)),
column(3))
),
column(8,  tabsetPanel(
      tabPanel("Simulation", plotOutput("plot1")), 
      tabPanel("Details", 
           withMathJax(
            helpText("MODEL:"),
            helpText("Species 1 $$\\frac{dN_1}{dt} = r_1 N_1 (\\frac{K_1-N_1-\\alpha_{12} N_2}{K_1})$$"),
            helpText("Species 2 $$\\frac{dN_2}{dt} = r_2 N_2 (\\frac{K_2-N_2-\\alpha_{21} N_1}{K_2})$$"),
          helpText("N_1-isocline $$N_2 = \\frac{K_1 - N_1}{\\alpha_{12}}$$"),
          helpText("N_2-isocline $$N_2 = K_2 - \\alpha_{21} N_1$$")
          )
           )
  )
))
)


# This creates the "behind the scenes" code (Server)
server = function(input, output) {
compLV=function(t, y, parameters){
   N1=y[1]
   N2=y[2]

   with(as.list(parameters),{
   dN1 = r1*N1*((K1-N1-a*N2)/K1)
   dN2 = r2*N2*((K2-N2-b*N1)/K2)
   res=c(dN1, dN2)
   list(res)
})
}

output$plot1 <- renderPlot({
N1star=(input$K1-input$a*input$K2)/(1-input$a*input$b)
N2star=(input$K2-input$b*input$K1)/(1-input$a*input$b)


times  = seq(0, input$Tmax, by=0.1)
parms=c(r1=input$r1, r2=input$r2,a=input$a,b=input$b,
K1=input$K1,K2=input$K2)
xstart = c(N1=input$N1, N2=input$N2)

out=ode(y=xstart,
   times=times,
   func=compLV,
   parms=parms)

   out=as.data.frame(out)


par(mfrow=c(1,2))  #This puts two plots side by side each other
plot(times, out$N1, ylab="abundance", xlab="time", type="l",
ylim=range(out[,2:3]))
lines(times, out$N2, col="red")
   legend("topright",
         legend=c("N1", "N2"),
         lty=c(1,1),
          col=c("black", "red"))

plot(NA,xlim=c(0,input$K1*2),ylim=c(0,input$K2*2), xlab="N1", ylab="N2")
fld=flowField(compLV, xlim=c(0,input$K1*2), ylim=c(0,input$K2*2), 
parameters=parms, system="two.dim", add=TRUE)
legend("topright",
         legend=c("N1-iso", "N2-iso", "Trajectory"),
         lty=c(1,1, 1),
          col=c("black", "red", "black"))

#null clines
curve((input$K1-x)/input$a,col="black",add=TRUE)
curve(input$K2-input$b*x,col="red",add=TRUE)
points(input$K1,0,pch=19)
points(0,input$K2,pch=19)
if(!any(c(N1star, N2star)<0)) points(N1star,N2star,pch=19)
lines(out[,2], out[,3], lwd=2)
    })
   }

shinyApp(ui, server, options = list(height =530))
```

Reference: