---
title: "LPA tribolium model"
author: Ottar N. Bjornstad
output: html_document
runtime: shiny
---

Version 0.5 June 28, 2019 
https://github.com/objornstad/

This Rmarkdown of the LPA tribolium was written by Ottar N. Bjornstad and is released with a CC-BY-NC license for anyone to improve and re-share (acknowledging origin). Please email me a copy if you make an update (onb1 at psu dot edu).

The app requires the shiny and sctterplot3d packages to be installed to run. 
```{r}
require(shiny)
require(scatterplot3d)
```

The LPA model (Costantino et al 1997) studies the stage-structured cannibalistic interactions among eggs, larvae (L), pupae (P) and adult (A) Tribolium flour beetles. The model uses a difference equation framework with a 2-week time step. basic equations are:
\begin{align}
L_{t+1} & = \underbrace{b A_t}_{\mbox{eggs}} \underbrace{e^{-c_{ea} A_t} e^{-c_{el}L_t}}_{\mbox{not cannibalized}}\\
P_{t+1} & =   L_t  \underbrace{(1-m_l) }_{\mbox{not died}}\\
A_{t+1} & =   P_t  \underbrace{e^{-c_{pa} A_t}}_{\mbox{not cannibalized}} + \underbrace{(1-m_a) A_t}_{\mbox{not died}}
\end{align}

Adults are assumed to lay $b$ eggs of which a fraction $1-e^{-c_{ea} A_t}$ is cannibalized by adults and $1-e^{-c_{el}L_t}$ by larvae; $c_{ea}$ and $c_{el}$ are cannibalism coefficients. Larvae suffers density independent mortality with probability $m_l$ per biweek. The egg and the larval stages are assumed to be two weeks. A fraction $1-e^{-c_{pa} A_t}$ of the pupae are cannibalized by adults. Adults suffer density independent mortality $m_a$ per biweek.

Costantino et al. (1997) used replacement experiments to increase adult mortality from the control of $m_l = 0.0076$ to $m_l = 0.96$ and manipulated pupal cannibalism rates to $c_{pa}$ = 0 (stable), 0.05 (8-cycle), 0.010 (invariant loop), 0.25 (chaotic), 0.35 (chaotic), 0.5 (3-cycle or chaotic), 1 (3-cycle)$.

Unchecking the LINES BOX in the app allows for better visualization of complex attractors in the 2D and 3D phase planes.

The  shiny app:

```{r, echo=FALSE}
ui = pageWithSidebar(
  tags$head(tags$style(
    HTML('
         #sidebar1 {
            background-color: #ECECEC;
        }
    
    #sidebar2 {
      background-color: #ECECEC
    }')
  )),
titlePanel(""),
fluidRow(
column(4, id = "sidebar2",
fluidRow(column(5, id = "sidebar1",
sliderInput("b", "b:", 6.598,
              min = 0, max = 10, step=0.1),
sliderInput("cel", "cel:", 1.209e-2,
              min = 0, max = 1, step=0.01),
sliderInput("cea", "cea:", 1.155e-2,
              min = 0, max = 1, step=0.01)
),
column(5, offset = 1, id = "sidebar1",
sliderInput("cpa", "cpa:", 1,
              min = 0, max = 1, step=0.01),
sliderInput("mua", "ma:", 0.96,
              min = 0, max = 1, step=0.01),
sliderInput("mul", "ml:", 0.2055,
              min = 0, max = 1, step=0.01)
#numericInput("L0", "initial L:", 50,
#              min = 0, max = 100),
#numericInput("P0", "initial P:", 0,
#              min = 0, max = 100),
#numericInput("A0", "initial A:", 0,
#              min = 0, max = 100)
),
column(1)),
fluidRow(
column(6, offset = 3, id = "sidebar1",
sliderInput("T", "Time range:",
                  min = 1, max = 500, value = c(100,150)),
checkboxInput("li", "lines", TRUE)),
column(3))
),
column(8, 
mainPanel(
  tabsetPanel(
      tabPanel("Time", plotOutput("plot1")), 
      tabPanel("2D Phase plane", plotOutput("plot2")),   
      tabPanel("3D Phase plane", plotOutput("plot3"))
      )
)))
)

# This creates the 'behind the scenes' code (Server)
server <- function(input, output) {

lpa = function(b=6.598, cel=1.209e-2, cea=1.155e-2, mul=0.2055, 
      cpa=4.7e-3, mua=7.629e-3, init=c(50,0,0), T=500){
#Initiate empty matrix to hold simulation
res = matrix(NA, nrow=500, ncol=3)

#Add column names
dimnames(res) = list(NULL, c("L", "P", "A"))

#store initial conditions in first row
res[1,] = init

for(i in 2:T){
     #Larval equation
     res[i,1] = b*res[i-1,3]*exp(-cel*res[i-1,1]-cea*res[i-1,3])
     #Pupal equation
     res[i,2] = res[i-1,1]*(1-mul)
     #Adult equation
     res[i,3] = res[i-1,2]*exp(-cpa*res[i-1,3])+res[i-1,3]*(1-mua)
     }
return(res)
}

output$plot1 <- renderPlot({
out=lpa(b=input$b, cel=input$cel, cea=input$cea, mul=input$mul,
  cpa=input$cpa, mua=input$mua)
 
 
plot(input$T[1]:input$T[2], out[input$T[1]:input$T[2],3], type = "b", xlab = "Week", 
  ylab = "Abundance", xlim=input$T, ylim=c(0, max(out[input$T[1]:input$T[2],c(1,3)])))
lines(input$T[1]:input$T[2], out[input$T[1]:input$T[2],1], col=2)
 legend("topleft",
        legend=c("A", "L"),
        lty=c(1,1),
        pch=c(1,NA),
         col=c("black", "red"))

   })

output$plot2 <- renderPlot({
out=lpa(b=input$b, cel=input$cel, cea=input$cea, mul=input$mul,
  cpa=input$cpa, mua=input$mua)
 
 
plot(out[input$T[1]:input$T[2],1], out[input$T[1]:input$T[2],3], type = ifelse(input$li==TRUE, "b", "p"), xlab = "L", 
  ylab = "A", xlim=c(0, max(out[input$T[1]:input$T[2],1])), 
  ylim=c(0, max(out[input$T[1]:input$T[2],3])))

   })

output$plot3 <- renderPlot({
out=lpa(b=input$b, cel=input$cel, cea=input$cea, mul=input$mul,
  cpa=input$cpa, mua=input$mua)
 
scatterplot3d(x=out[input$T[1]:input$T[2],1], y=out[input$T[1]:input$T[2],2], 
  z=out[input$T[1]:input$T[2],3], type = ifelse(input$li==TRUE, "b", "p"), xlim=c(0, max(out[input$T[1]:input$T[2],1])),
   ylim=c(0, max(out[input$T[1]:input$T[2],2])),
    zlim=c(0, max(out[input$T[1]:input$T[2],3])),
    xlab="L", ylab="P", zlab="A")

   })


  }

shinyApp(ui, server)
```

Reference:

Costantino RF, Desharnais RA, Cushing JM, Dennis B (1997) Chaotic dynamics 
            in an insect population. Science 275: 389-391