---
title: "Outbreak Response vaccination Desicion tool"
author: "Ottar N. Bjornstad"
output: html_document
runtime: shiny
---

```{r, include=FALSE}
using<-function(...) {
    libs<-unlist(list(...))
    req<-unlist(lapply(libs,require,character.only=TRUE))
    need<-libs[req==FALSE]
    if(length(need)>0){ 
        install.packages(need)
        lapply(need,require,character.only=TRUE)
    }
}

using("shiny", "deSolve")
```

Version 0.5-8 August 27, 2022 
https://github.com/objornstad/

This Rmarkdown of Outbreak Response Vaccination impact was written by Ottar N. Bjørnstad and Matt Ferrari and is released with a CC-BY-NC lisence for anyone to improve and re-share (acknowledging origin). Please email me a copy of any update (onb1 at psu dot edu). The app was originally developed as part of the epimdr-package (https://cran.r-project.org/package=epimdr; Bjørnstad 2019) and was motivated by the study of Grais et al. (2008).

The app requires the shiny and deSolve packages to be installed to run. "Run document" in Rstudio will launch the App.

The App shows predicted reduction in epidemic size for an Outbreak Response Vaccination Campaign in the face of an SIR epidemicwith sensitivity analyses with respect to date of start and various disease parameters.


```{r, echo=FALSE}
#' Launch a shiny-app to study outbreak-response vaccination campaigns
#' @details
#' Launch app for details
#' @examples
#' \dontrun{orv.app}
#' @export
ui=navbarPage("ORV",
   tabPanel("Retrospective analysis",
    sidebarLayout(
    sidebarPanel(
numericInput("day5", "Start day:", 60,
              min = 10, max = 3*365),
sliderInput("vaccine_target5", "target:", 0.7,
              min = 0, max = 1),
sliderInput("vaccine_efficacy5", "efficacy:", 0.9,
              min = 0, max = 1),
numericInput("intervention_length5", "duration:", 14,
              min = 1, max = 100),
sliderInput("R5", "R", 
              min = 1, max = 20, value=4),
numericInput("N5", "N:", 1E5,
              min = 1E2, max = 1E6),
numericInput("mtime5", "endtime:", 120,
              min = 10, max = 3*365),
    sliderInput("IP5", "Infectious period (days)", 5,
              min = 1, max = 100),
    sliderInput("LP5", "Latent period (days):", 8,
              min = 1, max = 100)),
    mainPanel(plotOutput("plot5"))
  )),
  tabPanel("Intervention day",
    sidebarLayout(
    sidebarPanel(
sliderInput("vaccine_target1", "target:", 0.7,
              min = 0, max = 1),
sliderInput("vaccine_efficacy1", "efficacy:", 0.9,
              min = 0, max = 1),
numericInput("intervention_length1", "duration:", 14,
              min = 1, max = 100),
numericInput("N1", "N:", 1E5,
              min = 1E2, max = 1E6),
numericInput("mtime1", "endtime:", 120,
              min = 10, max = 3*365),
    sliderInput("R1", "R", 
              min = 1, max = 20, value=4),
    sliderInput("IP1", "Infectious period (days)", 5,
              min = 1, max = 100),
    sliderInput("LP1", "Latent period (days):", 8,
              min = 1, max = 100)),
    mainPanel(plotOutput("plot1"))
  )),
   tabPanel("R sensitivity",
    sidebarLayout(
    sidebarPanel(
    sliderInput("R2", "R", 
              min = 1, max = 20, value=4),
numericInput("pm2", "+/-:", 0.5,
              min = 1, max = 10),
sliderInput("vaccine_target2", "target:", 0.7,
              min = 0, max = 1),
sliderInput("vaccine_efficacy2", "efficacy:", 0.9,
              min = 0, max = 1),
numericInput("intervention_length2", "duration:", 14,
              min = 1, max = 100),
numericInput("N2", "N:", 1E5,
              min = 1E2, max = 1E6),
numericInput("mtime2", "endtime:", 120,
              min = 10, max = 3*365),
    sliderInput("IP2", "Infectious period (days)", 5,
              min = 1, max = 100),
    sliderInput("LP2", "Latent period (days):", 8,
              min = 1, max = 100)),
    mainPanel(plotOutput("plot2"))
  )),
   tabPanel("Duration sensitivity",
    sidebarLayout(
    sidebarPanel(
numericInput("intervention_length3", "duration:", 14,
              min = 1, max = 100),
numericInput("pm3", "+/-:", 7,
              min = 1, max = 21),
sliderInput("vaccine_target3", "target:", 0.7,
              min = 0, max = 1),
sliderInput("vaccine_efficacy3", "efficacy:", 0.9,
              min = 0, max = 1),
    sliderInput("R3", "R", 
              min = 1, max = 20, value=4),
numericInput("N3", "N:", 1E5,
              min = 1E2, max = 1E6),
numericInput("mtime3", "endtime:", 120,
              min = 10, max = 3*365),
    sliderInput("IP3", "Infectious period (days)", 5,
              min = 1, max = 100),
    sliderInput("LP3", "Latent period (days):", 8,
              min = 1, max = 100)),
    mainPanel(plotOutput("plot3"))
  )),
   tabPanel("Cover sensitivty",
    sidebarLayout(
    sidebarPanel(
sliderInput("vaccine_target4", "target:", 0.7,
              min = 0, max = 1),
numericInput("pm4", "+/-:", 0.1,
              min = 0, max = .9),
sliderInput("vaccine_efficacy4", "efficacy:", 0.9,
              min = 0, max = 1),
numericInput("intervention_length4", "duration:", 14,
              min = 1, max = 100),
    sliderInput("R4", "R", 
              min = 1, max = 20, value=4),
numericInput("N4", "N:", 1E5,
              min = 1E2, max = 1E6),
numericInput("mtime4", "endtime:", 120,
              min = 10, max = 3*365),
    sliderInput("IP4", "Infectious period (days)", 5,
              min = 1, max = 100),
    sliderInput("LP4", "Latent period (days):", 8,
              min = 1, max = 100)),
    mainPanel(plotOutput("plot4"))
  )),
  tabPanel("Summary")
)

server=function(input, output){
######################################################
#SEIR model 
######################################################
simod<-function(t,x,parms){
################
#Parameters
#B = transmission rate
#1/r = latent period
#1/g = infectious period
#q = vaccine efficacy
#
#P = target coverage
#Dt = length of vaccination campaign
#T = Day of campaign start
#USAGE:
# times<-1:100
# xstrt<-c(S=.999,E=0,I=.001,R=0,K=0)
# par<-c(B=.5, r=1/7, g = 1/7, q = .8, P = 0, Dt = 10, T = 80)
# out<-as.data.frame(lsoda(xstrt,times,simod,par))
# plot(out$time,out$I,type="l")
#
#
# par<-c(B=.5, r=1/7, g = 1/7, q = .8, P = .99, Dt = 10, T = 50)
# out<-as.data.frame(lsoda(xstrt,times,simod,par))
# lines(out$time,out$I,col="red")

    S<-x[1]
    E<-x[2]
    I<-x[3]
    R<-x[4]
    K<-x[5]
    #
    with(as.list(parms),{
      Q<- ifelse(t<T | t>T+Dt,0,(-log(1-P)/Dt))
      dS<- -B*S*I-q*Q*S
      dE<- B*S*I-r*E
      dI<- r*E - g*I
      dR<- g*I+q*Q*S
      dK<-r*E
      res<-c(dS,dE,dI,dR,dK)
      list(res)
    })
  }
######################################################
  
#####################################################
#% Intervention
#####################################################
p_red<-function(R,vaccine_efficacy,target_vaccination,intervention_length, mtime=120, LP=7, IP=7, N=10000, step=1){
  steps<-(0:mtime)[seq(1,mtime,by=step)]
  p_red<-rep(NA,length(steps))
  xstrt<-c(S=1-1/N,E=0,I=1/N,R=0,K=0)   #starting values
  beta<- R/IP         #transmission rate
  t<-1
  for(i in 1:length(steps)){
    par<-c(B=beta, r=1/LP, g = 1/IP, q = vaccine_efficacy,
                        P = target_vaccination, Dt = intervention_length, T = steps[i])
    out<-as.data.frame(lsoda(xstrt,steps,simod,par))
    p_red[t]<-out$K[dim(out)[1]]
    t<-t+1
    cat("step ", i,"of ",floor(mtime/step), ".\r")
  }
  par<-c(B=beta, r=1/LP, g = 1/IP, q = vaccine_efficacy,
                P = 0, Dt = 0, T = Inf)
  outv<-as.data.frame(lsoda(xstrt,steps,simod,par))
        #fs should really be the prediction with steps=Inf?
  fs<-max(out$K)
  res<-list(out=cbind(steps,p_red/max(p_red)),
                  R=R,
                  vaccine_efficacy=vaccine_efficacy,
                  target_vaccination=target_vaccination,
                  intervention_length=intervention_length,
                  mtime=mtime, LP=LP, IP=IP, N=N, step=step,
                  virgin=outv$I, vfs=fs)
  class(res)<-"p_red"
  return(res)
}
######################################################

#####################################################
#plotting pred objects
#####################################################
plot.p_red<-function(object){
      plot(object$out[,1],object$out[,2],type="l", xlab="First intervention day", ylab="% final epidemic", ylim=c(0,1))
      title(paste("target= ", round(100*object$target_vaccination,0), "% campaign = ", object$intervention_length,"d"))
par(new=TRUE)
      plot(object$out[,1], object$virgin, col='red', axes=FALSE, xlab="", ylab="", type="l")
      legend(x="topleft", legend=c("natural epidemic", "%final size"), col=c("red", "black"), lty=c(1,1))
}

#EX
#out<-p_red(R=4,vaccine_efficacy=.9,target_vaccination=.5,intervention_length=14, step=2)
#plot(out)

#####################################################
#% Sensitivity analysis on R
#####################################################
R_compare<-function(R=c(2,4,8),vaccine_efficacy=.9,target_vaccination, intervention_length,mtime=120, LP=7, IP=7, N=10000, step=7){
  out<-numeric(0)
  for(j in 1:length(R)){
    tmp<-p_red(R=R[j],vaccine_efficacy=vaccine_efficacy,
                target_vaccination,intervention_length, mtime=mtime, LP=LP, IP=IP, N=N, step=step)
    out<-cbind(out,tmp$out[,2])
  }
  res<-list(R=R, p_red=out, T=tmp$out[,1],
                 vaccine_efficacy=vaccine_efficacy,
                 target_vaccination=target_vaccination,
                 intervention_length=intervention_length,
                 mtime=mtime, LP=LP, IP=IP, N=N, step=step)
  class(res)<-"Rcomp"
  return(res)
}

#####################################################
#plotting pred objects
#####################################################
plot.Rcomp<-function(object){
        plot(NA,xlim=range(object$T),ylim=c(0,1), xlab="First intervention day", ylab="%, final epidemic")
        title(paste("% final size: target= ", round(100*object$target_vaccination,0), "% campaign = ", object$intervention_length,"d"))
  for(j in 1:length(object$R)){
    lines(object$T,object$p_red[,j],lty=j)
  }
      legend(x="right", legend=c(object$R), lty=1:length(object$R), title="R=")

}

#EX
#res<-R_compare(R=c(1.5, 2.5, 3.5), vaccine_efficacy=.9,target_vaccination=.5,intervention_length=7)
#plot(res)

#####################################################
#% Sensitivity analysis on length of intervention
#####################################################
Int_compare<-function(R,vaccine_efficacy,target_vaccination,intervention_length=c(7,10,14),mtime=120, LP=7, IP=7, N=10000, step=7){
  out<-numeric(0)
  for(j in 1:length(intervention_length)){
    tmp<-p_red(R=R,vaccine_efficacy=vaccine_efficacy,
                target_vaccination,intervention_length[j], mtime=mtime, LP=LP, IP=IP, N=N, step=step)
    out<-cbind(out,tmp$out[,2])
  }
  res<-list(R=R, p_red=out, T=tmp$out[,1],
                 vaccine_efficacy=vaccine_efficacy,
                 target_vaccination=target_vaccination,
                 intervention_length=intervention_length,
                 mtime=mtime, LP=LP, IP=IP, N=N, step=step)
  class(res)<-"Intcomp"
  return(res)
}

#####################################################
#plotting Intcomp objects
#####################################################
plot.Intcomp<-function(object){
        plot(NA,xlim=range(object$T),ylim=c(0,1), xlab="First intervention day", ylab="% final epidemic")
        title(paste("% final size: target= ", round(100*object$target_vaccination,0), "%, R = ", object$R))
  for(j in 1:length(object$intervention_length)){
    lines(object$T,object$p_red[,j],lty=j)
  }
      legend(x="right", legend=c(object$intervention_length), lty=1:length(object$intervention_length), title="Campaign D:")

}

#EX
#res<-Int_compare(intervention_length=c(7,10,14),R=3, vaccine_efficacy=.9, target_vaccination=.5)
#plot(res)

#####################################################
#% Sensitivity analysis on target Vaccination
#####################################################
Vacc_compare<-function(R,vaccine_efficacy,target_vaccination=c(.50,.70,.90),intervention_length=7,mtime=120, LP=7, IP=7, N=10000, step=7){
  out<-numeric(0)
  for(j in 1:length(target_vaccination)){
    tmp<-p_red(R=R,vaccine_efficacy=vaccine_efficacy,
                target_vaccination[j],intervention_length, mtime=mtime, LP=LP, IP=IP, N=N, step=step)
    out<-cbind(out,tmp$out[,2])
  }
  res<-list(R=R, p_red=out, T=tmp$out[,1],
                 vaccine_efficacy=vaccine_efficacy,
                 target_vaccination=target_vaccination,
                 intervention_length=intervention_length,
                 mtime=mtime, LP=LP, IP=IP, N=N, step=step)
  class(res)<-"Vacccomp"
  return(res)
}

#####################################################
#plotting Vaccomp objects
#####################################################
plot.Vacccomp<-function(object){
        plot(NA,xlim=range(object$T),ylim=c(0,1), xlab="First intervention day", ylab="% final epidemic")
        title(paste("% final size: R = ", object$R))
  for(j in 1:length(object$target_vaccination)){
    lines(object$T,object$p_red[,j],lty=j)
  }
      legend(x="right", legend=c(round(100*object$target_vaccination,0)), lty=1:length(object$target_vaccination), title="Target %")
}

#EX
#res<-Vacc_compare(target_vaccination=c(.50,.70,.90), R=4, vaccine_efficacy=.9,intervention_length=7)
#plot(res)

#####################################################
#Mortality analysis on R
#####################################################
M_red<-function(object, case_fatality){
        par(mfrow=c(2,1))
        plot(object)
  out<-(1-object$p_red)*object$N*case_fatality
  plot(NA,xlim=c(1,max(object$T)),ylim=c(0,max(out)), xlab="First intervention day", ylab="Extra survivors")
  for(j in 1:dim(out)[2]){
    lines(object$T, out[,j], lty=j)
  }
  title(paste("Reduced burden of mortality; N = ", object$N))
        par(mfrow=c(1,1))
}

#EX
#res<-Vacc_compare(target_vaccination=c(.50,.70,.90), R=4, vaccine_efficacy=.9,intervention_length=7)
#M_red(res, 0.5)


#####################################################
#Retrospective
#####################################################
retro<-function(R,day, vaccine_efficacy,target_vaccination,intervention_length, mtime=120, LP=7, IP=7, N=10000){
  steps<-1:mtime
        out<-matrix(NA,nrow=mtime, ncol=3)
  xstrt<-c(S=1-1/N,E=0,I=1/N,R=0,K=0)   #starting values
  beta<- R/IP         #transmission rate
  t<-1
  par<-c(B=beta, r=1/LP, g = 1/IP, q = vaccine_efficacy,
                P = 0, Dt = 0, T = Inf)
  outv<-as.data.frame(lsoda(xstrt,steps,simod,par))
        #fsv and fsi should really be with steps=Inf?
  fsv<-max(outv$K)

  par<-c(B=beta, r=1/LP, g = 1/IP, q = vaccine_efficacy,
             P = target_vaccination, Dt = intervention_length, T = day)

        outi<-as.data.frame(lsoda(xstrt,steps,simod,par))

  fsi<-max(outi$K)
        out[,1]<-steps
        out[,2]<-outv$I
        out[,3]<-outi$I
        res<-list(out=out,
                  redn=fsi/fsv,
                  R=R,
                  vaccine_efficacy=vaccine_efficacy,
                  target_vaccination=target_vaccination,
                  intervention_length=intervention_length,
                  mtime=mtime, LP=LP, IP=IP, N=N, day=day)
  class(res)<-"retro"
  return(res)
}

#####################################################
#plotting pred objects
#####################################################
plot.retro<-function(object){
      plot(object$out[,1],object$out[,2],type="l", ylim=c(0,max(object$out[,2])), xlab='day', ylab='prevalence')
      polygon(c(object$day, object$day, object$day+object$intervention_length,
               object$day+object$intervention_length), c(-0.1,1,1,-.1), col="gray")
      lines(object$out[,1],object$out[,2])
      lines(object$out[,1], object$out[,3], col='red')
      title(paste("final size: ", round(100*(object$redn),1), "% (R=",
               object$R,", target=", 100*object$target_vaccination,"%)", sep=""))
      legend(x="topright", legend=c("natural epidemic", "w intervention"),
               col=c("black", "red"), lty=c(1,1))
      text(x=object$day+object$intervention_length, y=0, pos=4,
                labels=paste(object$intervention_length,
                "d intervention from", object$day))
}


output$plot1 <- renderPlot({
out<-p_red(R=input$R1,input$vaccine_efficacy1,input$vaccine_target1,input$intervention_length1, input$mtime1, input$LP1, input$IP1, input$N1, step=1)
plot(out)}
)

output$plot2 <- renderPlot({
R2=c(input$R2-input$pm2, input$R2, input$R2+input$pm2)
R2[R2<0]=0
out2=R_compare(R=R2, input$vaccine_efficacy2,input$vaccine_target2,input$intervention_length2, 
  input$mtime2, input$LP2, input$IP2, input$N2, step=1)
plot(out2)}
)

output$plot3 <- renderPlot({
il=c(input$intervention_length3-input$pm3,
input$intervention_length3, input$intervention_length3+input$pm3)
il[il<0]=0
out3=Int_compare(R=input$R3, input$vaccine_efficacy3,input$vaccine_target3,il,  
  input$mtime3, input$LP3, input$IP3, input$N3, step=1)
plot(out3)}
)

output$plot4 <- renderPlot({
vt=c(input$vaccine_target4-input$pm4,
input$vaccine_target4, input$vaccine_target4+input$pm4)
vt[vt<0]=0
vt[vt>1]=1
out4=Vacc_compare(R=input$R4, input$vaccine_efficacy4,vt,input$intervention_length4, 
 input$mtime4, input$LP4, input$IP4, input$N4, step=1)
plot(out4)}
)

output$plot5 <- renderPlot({
out5<-retro(R=input$R5, day=input$day5, input$vaccine_efficacy5,input$vaccine_target5,
  input$intervention_length5, input$mtime5, input$LP5, input$IP5, input$N5)
plot(out5)}
)

}

shinyApp(ui, server, options = list(height = 1000))
```

REFERENCES:
Bjørnstad, O.N. (2018) Epidemics: Models and Data using R. Springer (312 pp) ISBN 978-3-319-97487-3 https://www.springer.com/gp/book/9783319974866

Grais, R. F., Conlan, A. J. K., Ferrari, M. J., Djibo, A., Le Menach, A., Bjørnstad,
O. N., et al. (2008). Time is of the essence: Exploring a measles outbreak response
vaccination in niamey, niger. Journal of the Royal Society Interface, 5(18), 67–74.