---
title: "Parasitoid-host models with embedded ShinyApps"
author: Ottar N. Bjørnstad
output: html_document
runtime: shiny
---

Version 0.5-8 June 28, 2022 
https://github.com/objornstad/

This Rmarkdown of Nicholson-Bailey (Nicholson and Bailey 1935) and negative-binomial (May 1978) parasitoid-host model was written by Ottar N. Bjørnstad and is released with a CC-BY-NC license for anyone to improve and re-share (acknowledging origin). Please email me a copy if you make an update (onb1 at psu dot edu). The app was originally developed as part of the epimdr-package (https://cran.r-project.org/package=epimdr; Bjørnstad 2019).

The app requires the shiny package to be installed to run. "Run document" in Rstudio will launch the App.

```{r, include=FALSE}
require(shiny)
```

The Nicholson-Bailey (1935) model assumes a population of randomly searching parasitoids, $P$, parasitizing immature stages of their host, $H$. The host growth rate is $R$ and the parasitoid search efficiency is $a$. Given the random search, the fraction of hosts escaping parasitization is $\mbox{exp}(- a P)$ leading to two equations: 

$\begin{aligned}
H_{t+1} & = \underbrace{R H_t}_{\mbox{reproduction}} \underbrace{e^{-a P_t}}_{\mbox{not parasitized}}\\
P_{t+1} & =   R H_t  \underbrace{(1-e^{-a P_t}) }_{\mbox{parasitized}}
\end{aligned}$

Nicholson and Bailey (1935) showed that the equilibrium $\{H^* = \frac{\mbox{log}(R)}{a (R-1)}, P^* = \frac{\mbox{log}(R)}{a}\}$" is an unstable focus leading to increasingly violent consumer-resource cycles.

The Nicholson-Bailey shiny app is:

```{r, echo=FALSE}
# This creates the NB User Interface (UI)
ui = pageWithSidebar(
headerPanel("Nicholson-Bailey Model"),
sidebarPanel(
sliderInput("R", "Growth rate (R):", 1.1,
              min = 1, max = 2, step=.01),
sliderInput("a", "Search efficiency (a):", 0.03,
              min = 0, max = .5),
numericInput("P0", "Initial parasitoid:", 10,
              min = 1, max = 100),
numericInput("H0", "Initial host:", 20,
              min = 1, max = 100),
numericInput("Tmax", "Tmax:", 100,
              min = 1, max = 500)
),
mainPanel(tabsetPanel(
  tabPanel("Simulation", plotOutput("plot1", height = 500)),
  tabPanel("Phase plane", plotOutput("plot2", height = 500)),
  tabPanel("Details",
    withMathJax(
            helpText("MODEL:"),
          helpText("Host $$H_t = R H_{t-1} (1 - \\mbox{exp}(- a P_{t-1}))$$"),
          helpText("Parasitoid $$P_t = R H_{t-1} \\mbox{exp}(- a P_{t-1})$$"),
          helpText("Equilibria $$H^* = \\frac{\\mbox{log}(R)}{a (R-1)}, 
            P^* = \\frac{\\mbox{log}(R)}{a}$$")),
          helpText("REFERENCE: Nicholson AJ, Bailey VA (1935) The balance of animal populations. 
            Proceedings of the Zoological Society of London 3: 551-598")
))
)
)

# This creates the 'behind the scenes' code (Server)
server = function(input, output) {
 NB = function(R, a, T = 100, H0 = 10, P0 = 1){
   #T is length of simulation (number of time-steps)
   #H0 and P0 are initial numbers
   #we provide default parameters except for R and a

   H=rep(NA, T) #host series
   P=rep(NA, T) #parasitoid series

   H[1] = H0 #Initiating the host series
   P[1] = P0 #Initiating the host series

   for(t in 2:T){
     H[t] = R * H[t-1] * exp(- a * P[t-1])
     P[t] = R * H[t-1] * (1-exp(- a * P[t-1]))
     if(P[t-1]==0) break
   } #end of loop

   #the two vectors of results are stored in a "list"
   res= data.frame(H = H, P = P)
 
   #the list is passed out of this function
   return(res)
} #end of function



  output$plot1 <- renderPlot({

    sim= NB(R=input$R, a=input$a, H0=input$H0, P0=input$P0, T=input$Tmax)
    time = 1:input$Tmax

    plot(time, sim$H, type= "b",xlab = "Generations", ylab = "Abundance", 
      ylim = range(sim, na.rm=TRUE))
    points(time, sim$P, type = "b", col = "red")
     legend("topleft",
        legend=c("H", "P"),
        lty=c(1,1),
        pch=c(1,1),
        col=c("black", "red"))
   })


  output$plot2 <- renderPlot({

    sim= NB(R=input$R, a=input$a, H0=input$H0, P0=input$P0, T=input$Tmax)
    time = 1:input$Tmax
    Hstar=log(input$R)/(input$a*(input$R-1))
    Pstar=log(input$R)/input$a
    plot(sim$H, sim$P, type= "b",xlab = "Host", ylab = "Parasitoid")
    points(Hstar, Pstar, col=2, pch=19)
   })

}
  
shinyApp(ui, server, options = list(height = 640))
```

May (1978) proposed that heterogeneities in searching can stabilize parasitoid host dynamics. Assuming a negative binomial (rather than Poisson) attack probability the model is:

$\begin{aligned}
H_{t+1} & = R H_t (1+\frac{a P_t}{k})^{-k}\\
P_{t+1} & = R H_t (1-(1+\frac{a P_t}{k})^{-k}),
\end{aligned}$

where $k$ is the aggregation parameter ($k \rightarrow \infty$, Neg Bin $\rightarrow$ Poisson). He coined the CV$^2$-rule which says that if the coefficient-of-variation in attack rate is greater than 1, the parasitoid-host dynamics stabilizes; The CV for the negative binomial being $1/\sqrt{k}$. The equilibrium $P^* = k (R^{1/k}-1) /a, H^*=P^* R / (R-1)$ is either stable or unstable depending on whether k is smaller or greater than 1.


```{r, echo=FALSE}
# This creates the User Interface (UI)
ui = pageWithSidebar(
headerPanel("May's Parasitoid-host Model"),
sidebarPanel(
sliderInput("R", "Growth rate (R):", 1.1,
              min = 1, max = 2, step=.01),
sliderInput("a", "Search efficiency (a):", 0.1,
              min = 0, max = .5),
sliderInput("k", "aggregation (k):", 1.5,
              min = 0.1, max = 3, step=0.1),
numericInput("P0", "Initial parasitoid:", 10,
              min = 1, max = 100),
numericInput("H0", "Initial host:", 20,
              min = 1, max = 100),
numericInput("Tmax", "Tmax:", 100,
              min = 1, max = 500)
),
mainPanel(tabsetPanel(
  tabPanel("Simulation", plotOutput("plot1", height = 500)),
  tabPanel("Phase plane", plotOutput("plot2", height = 500)),
     tabPanel("Details", 
    withMathJax(
         helpText("MODEL:"),
             helpText("Host $$H_t = R H_{t-1} (1 + a P_{t-1})^k$$"),
          helpText("Parasitoid $$P_t = R H_{t-1} (1-(1 + a P_{t-1})^k)$$"),
          helpText("REFERENCE: May RM (1978) Host-parasitoid systems in patchy 
            environments: a phenomenological model. J Anim Ecol 47: 833-843")
)
)
)
)
)
server = function(input, output) {
 NB = function(R, a, k, T = 100, H0 = 10, P0 = 1){
   #T is length of simulation (number of time-steps)
   #H0 and P0 are initial numbers
   #we provide default parameters except for R and a

   H=rep(NA, T) #host series
   P=rep(NA, T) #parasitoid series

   H[1] = H0 #Initiating the host series
   P[1] = P0 #Initiating the host series

   for(t in 2:T){
     H[t] = R * H[t-1] * (1+ a * P[t-1]/k)^(-k)
     P[t] = R * H[t-1] * (1-(1+ a * P[t-1]/k)^(-k))
     if(P[t-1]==0) break
   } #end of loop

   #the two vectors of results are stored in a "list"
   res= data.frame(H = H, P = P)
 
   #the list is passed out of this function
   return(res)
} #end of function



  output$plot1 <- renderPlot({

    sim= NB(R=input$R, a=input$a, k=input$k, H0=input$H0, P0=input$P0, T=input$Tmax)
    time = 1:input$Tmax

    plot(time, sim$H, type= "b",xlab = "Generations", ylab = "Abundance", 
      ylim = range(sim, na.rm=TRUE))
    points(time, sim$P, type = "b", col = "red")
     legend("topleft",
        legend=c("H", "P"),
        lty=c(1,1),
        pch=c(1,1),
        col=c("black", "red"))
   })

output$plot2 <- renderPlot({

    sim= NB(R=input$R, a=input$a, k=input$k, H0=input$H0, P0=input$P0, T=input$Tmax)
    time = 1:input$Tmax
   
    plot(sim$H, sim$P, type= "b",xlab = "Host", ylab = "Parasitoid")
    #Pstar=input$k*(input$R^(1/input$k)-1)/input$a
    #Hstar=Pstar*input$R/(input$R-1)
    #points(Hstar, Pstar, col=2, pch=19)
   })

}  
shinyApp(ui, server, options = list(height = 660))
```

References:

Bjørnstad, O.N. (2018) Epidemics: models and data using R. Springer.

May, R.M. (1978) Host-parasitoid systems in patchy environments: a phenomenological model. J Anim Ecol 47: 833-843

Nicholson A.J. and Bailey V.A. (1935) The balance of animal populations. Proceedings of the Zoological Society of London 3: 551-598.