---
title: "Rosenzweig-MacArthur predator-prey model"
author: Ottar N. Bjørnstad
output: html_document
runtime: shiny
---

Version 0.5.4 February 4, 2020 
https://github.com/objornstad

This Rmarkdown of the Rosenzweig-MacArthur predator-prey model was written by Ottar N. Bjørnstad and is released with a CC-BY-NC lisence for anyone to improve and re-share (acknowledging origin). Please email me a copy of update (onb1 at psu dot edu).

The app requires the shiny and deSolve packages to be installed to run. 

```{r, echo=FALSE, include=FALSE}
require(shiny)
require(deSolve)
require(phaseR)
```

The basic equations for the consumer-resource interaction between prey (N)  and predators (P) are:
\begin{align}
    \frac{dN}{dt} =& \underbrace{r N (\frac{K-N}{K})}_{\mbox{N growth}} - \underbrace{\frac{a N P}{c + N}}_{\mbox{predation}}\\
     \frac{dP}{dt} =& \underbrace{\frac{b N P}{c + N}}_{\mbox{P growth}} - \underbrace{g P}_{\mbox{P death}}
\end{align}

Prey ($N$) are assumed to grow acording to the logistic model with a maximum growth rate, $r$ and carrying-capacity, $K$. Predators ($P$)are feeding according to a Type-II functional respose with a maximum search efficiency, $a$ and half-saturation constant $c$. Predators have a conversion efficiency of $b/a$ and a life-expectancy of $1/g$.

The isoclines (sometimes called the nullclines) of this system are given by the solution to the 
equations $\frac{dN}{dt} = 0$ and $\frac{dP}{dt} = 0$ and partitions the phase plane into regions 
were $N$ and $P$ are increasing and decreasing. The $N$-isocline is $P^* = (r-rN/K)(c+N)/a$
and the P-isocline is $N^* = gc/(b-g)$. The equilibrium is
\begin{align}
N^* =& gc/(b-g) \\
P^* =& (r-rN^*/K)(c+N^*)/a
\end{align}
The  shiny app is:

```{r, echo=FALSE}
# This creates the User Interface (UI)
ui = fluidPage(
  tags$head(tags$style(
    HTML('
         #sidebar1 {
            background-color: #ECECEC;
        }
    
    #sidebar2 {
      background-color: #ECECEC
    }')
  )),
fluidRow(
column(4, id = "sidebar2",
fluidRow(column(5, id = "sidebar1",
sliderInput("r", "r:", 0.1,
               min = 0, max = 1, step=0.01),
sliderInput("K", "K:", 90,
               min = 0, max = 300, step=1),
sliderInput("a", "a:", 0.2,
               min = 0, max = 1, step=0.01),
numericInput("N", "initial N:", 10,
               min = 0, max = 100)),
column(5, offset = 1, id = "sidebar1",
sliderInput("c", "c:", 20,
               min = 0, max = 100, step=0.1),
sliderInput("b", "b:", 0.1,
               min = 0, max = 1, step=0.01),
sliderInput("g", "g:", 0.05,
               min = 0, max = 1, step=0.01),
numericInput("P", "initial P:", 1,
               min = 0, max = 100)),
column(1)),
fluidRow(
column(6, offset = 3, id = "sidebar1",
numericInput("Tmax", "Tmax:", 1000,
               min = 0, max = 5000)),
column(3))
),
#column(8, plotOutput("plot1", height = 500))
column(8,  tabsetPanel(
      tabPanel("Time", plotOutput("plot1")), 
      tabPanel("Phase plane", plotOutput("plot2")),
      tabPanel("Details", 
           withMathJax(
       helpText("MODEL:"),
            helpText("Prey $$\\frac{dN}{dt} = r N (1-\\frac{N}{K}) - \\frac{a N P}{c+N}$$"),
          helpText("Predator $$\\frac{dP}{dt} = \\frac{b N P}{c+N} - g P$$"),
          helpText("N-isocline $$P^* = (r-rN/K)(c+N)/a$$"),
          helpText("P-isocline $$N^* = gc/(b-g)$$"),
          helpText("Equilibria $$N^* = gc/(b-g), P^* = (r-rN^*/K)(c+N^*)/a$$"),
         helpText("REFERENCE: Rosenzweig ML, MacArthur RH (1963) Graphical representation 
              and stability conditions of predator-prey interactions. Am Nat 97: 209-223")
          ))

   
  )
)
)
)

# This creates the "behind the scenes" code (Server)
server = function(input, output){
RM=function(t, y, parameters){
  N=y[1]
  P=y[2]

  r=parameters["r"]
  K=parameters["K"]
  a=parameters["a"]
  c=parameters["c"]
  b=parameters["b"]
  g=parameters["g"]

    dN = r*N*(1-N/K)-a*N*P/(c+N)
    dP = b*N*P/(c+N)-g*P
    res=c(dN,dP)
    list(res)
}

output$plot1 <- renderPlot({

 times  = seq(0, input$Tmax, by=0.1)
 parms=c(r=input$r, K=input$K,a=input$a,
   c=input$c,b=input$b,g=input$g)
 xstart = c(N=input$N, P=input$P)

 out=ode(y=xstart,
    times=times,
    func=RM,
    parms=parms)

    out=as.data.frame(out)

  r=parms["r"]
  K=parms["K"]
  a=parms["a"]
  c=parms["c"]
  b=parms["b"]
  g=parms["g"]


 plot(out$time, out$N, ylab="abundance", xlab="time", type="l", ylim=range(out[,2:3]))
 lines(out$time, out$P, col="red")
    legend("topright",
          legend=c("N", "P"),
          lty=c(1,1),
           col=c("black", "red"))
})

output$plot2 <- renderPlot({

 times  = seq(0, input$Tmax, by=0.1)
 parms=c(r=input$r, K=input$K,a=input$a,
   c=input$c,b=input$b,g=input$g)
 xstart = c(N=input$N, P=input$P)

 out=ode(y=xstart,
    times=times,
    func=RM,
    parms=parms)

    out=as.data.frame(out)

  r=parms["r"]
  K=parms["K"]
  a=parms["a"]
  c=parms["c"]
  b=parms["b"]
  g=parms["g"]

#null clines
plot(out$N, out$P, ylab='predator', xlab='prey', type='l', 
xlim=range(out$N), ylim= range(out$P))
abline(h=0, col = "green")
abline(v=0, col = "red")
curve(r*(1-x/K)*(c+x)/a,from = 0, to = max(c(90, out$N)), col = "green",add=T)
abline(v=g*c/(b-g),col = "red")
fld=flowField(RM, xlim=range(out$N), ylim=range(out$P), 
parameters=parms, system="two.dim", add=TRUE)
   legend("topright",
          legend=c("N-iso", "P-iso"),
          lty=c(1,1),
           col=c("green", "red"))

# points(Nstar,Pstar,pch = 1)

})


}

shinyApp(ui, server, options = list(height = 500))
```

Reference:

Rosenzweig, M.L. and MacArthur, R.H. (1963) Graphical representation 
              and stability conditions of predator-prey interactions. Am Nat 97: 209-223
