---
title: "Rosenzweig-MacArthur predator-prey model"
author: Ottar N. Bjørnstad
output: html_document
runtime: shiny
---

Version 0.5.3 August 26, 2019 
https://github.com/objornstad

This Rmarkdown of the Rosenzweig-MacArthur predator-prey model was written by Ottar N. Bjørnstad and is released with a CC-BY-NC lisence for anyone to improve and re-share (acknowledging origin). Please email me a copy of update (onb1 at psu dot edu).

The app requires the shiny and deSolve packages to be installed to run. 

```{r echo=FALSE}
require(shiny)
require(deSolve)
#The flowField function was written by Michael J Grayling (https://github.com/mjg211) as part of 
#recently archived phaseR-package and is reproduced here under the original GPL-3 license
flowField=function (deriv, xlim, ylim, parameters = NULL, system = "two.dim", 
   points = 21, col = "gray", arrow.type = "equal", arrow.head = 0.05, 
    frac = 1, add = TRUE, xlab = "x", ylab = "y", state.names = c("x", 
        "y"), ...) 
{
    if ((!is.vector(xlim)) | (length(xlim) != 2)) {
        stop("xlim is not a vector of length 2 as required")
    }
    if (xlim[2] <= xlim[1]) {
        stop("xlim[2] is less than or equal to xlim[1]")
    }
    if ((!is.vector(ylim)) | (length(ylim) != 2)) {
        stop("ylim is not a vector of length 2 as required")
    }
    if (ylim[2] <= ylim[1]) {
        stop("ylim[2] is less than or equal to ylim[1]")
    }
    if (points <= 0) {
        stop("points is less than or equal to zero")
    }
    if (!(system %in% c("one.dim", "two.dim"))) {
        stop("system must either be set to \"one.dim\" or \"two.dim\"")
    }
    if (is.vector(col) == FALSE) {
        stop("col is not a vector as required")
    }
    if (length(col) > 1) {
        col <- col[1]
        print("Note: col has been reset as required")
    }
    if (!(arrow.type %in% c("proportional", "equal"))) {
        stop("arrow.type must either be set to \"proportional\" or \"equal\"")
    }
    if (arrow.head <= 0) {
        stop("arrow.head is less than or equal to zero")
    }
    if (frac <= 0) {
        stop("frac is less than or equal to zero")
    }
    if (!is.logical(add)) {
        stop("add must be logical")
    }
    x <- seq(from = xlim[1], to = xlim[2], length = points)
    y <- seq(from = ylim[1], to = ylim[2], length = points)
    dx <- matrix(0, ncol = points, nrow = points)
    dy <- matrix(0, ncol = points, nrow = points)
    xmax.length <- x[2] - x[1]
    ymax.length <- y[2] - y[1]
    if (add == FALSE) {
        plot(1, xlim = c(xlim[1] - xmax.length, xlim[2] + xmax.length), 
            ylim = c(ylim[1] - ymax.length, ylim[2] + ymax.length), 
            type = "n", xlab = xlab, ylab = ylab, ...)
    }
    if (system == "one.dim") {
        for (i in 1:points) {
            dy[1, i] <- deriv(0, setNames(c(y[i]), state.names[1]), 
                parameters)[[1]]
        }
        for (i in 2:points) {
            dy[i, ] <- dy[1, ]
        }
        abs.dy <- abs(dy)
        abs.dy.non <- abs.dy[which(abs.dy != 0)]
        max.abs.dy <- max(abs(dy))
        coefficient <- frac * min(xmax.length, ymax.length)/(2 * 
            sqrt(2) * max(sqrt(2 * abs.dy.non/(abs.dy.non + (1/abs.dy.non))), 
            sqrt(2 * (1/abs.dy.non)/(abs.dy.non + (1/abs.dy.non)))))
        for (i in 1:points) {
            for (j in 1:points) {
                if (dy[i, j] != 0) {
                  factor <- sqrt(2/(abs.dy[i, j] + (1/abs.dy[i, 
                    j])))
                  y.shift <- coefficient * factor * sqrt(abs.dy[i, 
                    j])
                  x.shift <- coefficient * factor/sqrt(abs.dy[i, 
                    j])
                  if (dy[i, j] < 0) {
                    y.shift <- -y.shift
                  }
                }
                if (dy[i, j] == 0) {
                  y.shift <- 0
                  x.shift <- coefficient * sqrt(2)
                }
                if (arrow.type == "proportional") {
                  if (dy[i, j] != 0) {
                    prop <- abs.dy[i, j]/max.abs.dy
                    y.shift <- y.shift * prop
                    x.shift <- x.shift * prop
                  }
                  if (dy[i, j] == 0) {
                    x.shift <- y.shift * mean(abs.dy)/max.abs.dy
                  }
                }
                arrows(x[i] - x.shift, y[j] - y.shift, x[i] + 
                  x.shift, y[j] + y.shift, length = arrow.head, 
                  col = col, ...)
            }
        }
        return(list(add = add, arrow.head = arrow.head, arrow.type = arrow.type, 
            col = col, deriv = deriv, dy = dy, frac = frac, parameters = parameters, 
            points = points, system = system, x = x, xlab = xlab, 
            xlim = xlim, y = y, ylab = ylab, ylim = ylim))
    }
    else {
        for (i in 1:length(x)) {
            for (j in 1:length(y)) {
                df <- deriv(0, setNames(c(x[i], y[j]), state.names), 
                  parameters)
                dx[i, j] <- df[[1]][1]
                dy[i, j] <- df[[1]][2]
            }
        }
        abs.dx <- abs(dx)
        abs.dy <- abs(dy)
        abs.dx.non <- abs.dx[which((abs.dx != 0) & (abs.dy != 
            0))]
        abs.dy.non <- abs.dy[which((abs.dx != 0) & (abs.dy != 
            0))]
        max.length <- max(sqrt(dx^2 + dy^2))
        coefficient <- frac * min(xmax.length, ymax.length)/(2 * 
            sqrt(2) * max(sqrt(2 * (abs.dy.non/abs.dx.non)/((abs.dy.non/abs.dx.non) + 
            (abs.dx.non/abs.dy.non))), sqrt(2 * (abs.dx.non/abs.dy.non)/((abs.dy.non/abs.dx.non) + 
            (abs.dx.non/abs.dy.non)))))
        for (i in 1:points) {
            for (j in 1:points) {
                if ((dx[i, j] != 0) | (dy[i, j] != 0)) {
                  if ((dx[i, j] != 0) & (dy[i, j] != 0)) {
                    factor <- sqrt(2/((abs.dy[i, j]/abs.dx[i, 
                      j]) + (abs.dx[i, j]/abs.dy[i, j])))
                    y.shift <- coefficient * factor * sqrt(abs.dy[i, 
                      j]/abs.dx[i, j])
                    x.shift <- coefficient * factor/sqrt(abs.dy[i, 
                      j]/abs.dx[i, j])
                    if (dy[i, j] < 0) {
                      y.shift <- -abs(y.shift)
                    }
                    if (dx[i, j] < 0) {
                      x.shift <- -abs(x.shift)
                    }
                  }
                  if ((dx[i, j] == 0) & (dy[i, j] != 0)) {
                    y.shift <- coefficient * sqrt(2)
                    x.shift <- 0
                    if (dy[i, j] < 0) {
                      y.shift <- -abs(y.shift)
                    }
                  }
                  if ((dx[i, j] != 0) & (dy[i, j] == 0)) {
                    y.shift <- 0
                    x.shift <- coefficient * sqrt(2)
                    if (dx[i, j] < 0) {
                      x.shift <- -abs(x.shift)
                    }
                  }
                  if (arrow.type == "proportional") {
                    prop <- sqrt((abs.dx[i, j]^2 + abs.dy[i, 
                      j]^2))/max.length
                    y.shift <- y.shift * prop
                    x.shift <- x.shift * prop
                  }
                  arrows(x[i] - x.shift, y[j] - y.shift, x[i] + 
                    x.shift, y[j] + y.shift, length = arrow.head, 
                    col = col, ...)
                }
            }
        }
    }
    return(list(add = add, arrow.head = arrow.head, arrow.type = arrow.type, 
        col = col, deriv = deriv, dx = dx, dy = dy, frac = frac, 
        parameters = parameters, points = points, system = system, 
        x = x, xlab = xlab, xlim = xlim, y = y, ylab = ylab, 
        ylim = ylim))
}
```

The basic equations for the consumer-resource interaction between prey (N)  and predators (P) are:
\begin{align}
    \frac{dN}{dt} =& \underbrace{r N (\frac{K-N}{K})}_{\mbox{N growth}} - \underbrace{\frac{a N P}{c + P}}_{\mbox{predation}}\\
     \frac{dP}{dt} =& \underbrace{\frac{b N P}{c + P}}_{\mbox{P growth}} - \underbrace{g P}_{\mbox{P death}}
\end{align}

Prey are assumed to grow acording to the logistic model with a maximum growth rate, $r$ and carrying-capacity, $K$. Predators are feeding according to a Type-II functional respose with a maximum search efficiency, $a$ and half-saturation constant $c$. Predators have a conversion efficiency of $b/a$ and a life-expectancy of $1/g$.

The isoclines (sometimes called the nullclines) of this system are given by the solution to the 
equations $\frac{dN}{dt} = 0$ and $\frac{dP}{dt} = 0$ and partitions the phase plane into regions 
were $N$ and $P$ are increasing and decreasing. The $N$-isocline is $P^* = (r-rN/K)(c+N)/a$
and the P-isocline is $N^* = gc/(b-g)$. The equilibrium is
\begin{align}
N^* =& gc/(b-g) \\
P^* =& (r-rN^*/K)(c+N^*)/a
\end{align}
The  shiny app is:

```{r, echo=FALSE}
# This creates the User Interface (UI)
ui = fluidPage(
  tags$head(tags$style(
    HTML('
         #sidebar1 {
            background-color: #ECECEC;
        }
    
    #sidebar2 {
      background-color: #ECECEC
    }')
  )),
fluidRow(
column(4, id = "sidebar2",
fluidRow(column(5, id = "sidebar1",
sliderInput("r", "r:", 0.1,
               min = 0, max = 1, step=0.01),
sliderInput("K", "K:", 90,
               min = 0, max = 300, step=1),
sliderInput("a", "a:", 0.2,
               min = 0, max = 1, step=0.01),
numericInput("N", "initial N:", 10,
               min = 0, max = 100)),
column(5, offset = 1, id = "sidebar1",
sliderInput("c", "c:", 20,
               min = 0, max = 100, step=0.1),
sliderInput("b", "b:", 0.1,
               min = 0, max = 1, step=0.01),
sliderInput("g", "g:", 0.05,
               min = 0, max = 1, step=0.01),
numericInput("P", "initial P:", 1,
               min = 0, max = 100)),
column(1)),
fluidRow(
column(6, offset = 3, id = "sidebar1",
numericInput("Tmax", "Tmax:", 1000,
               min = 0, max = 5000)),
column(3))
),
#column(8, plotOutput("plot1", height = 500))
column(8,  tabsetPanel(
      tabPanel("Time", plotOutput("plot1")), 
      tabPanel("Phase plane", plotOutput("plot2")),
      tabPanel("Details", 
           withMathJax(
       helpText("MODEL:"),
            helpText("Prey $$\\frac{dN}{dt} = r N (1-\\frac{N}{K}) - \\frac{a N P}{c+N}$$"),
          helpText("Predator $$\\frac{dP}{dt} = \\frac{b N P}{c+N} - g P$$"),
          helpText("N-isocline $$P^* = (r-rN/K)(c+N)/a$$"),
          helpText("P-isocline $$N^* = gc/(b-g)$$"),
          helpText("Equilibria $$N^* = gc/(b-g), P^* = (r-rN^*/K)(c+N^*)/a$$"),
         helpText("REFERENCE: Rosenzweig ML, MacArthur RH (1963) Graphical representation 
              and stability conditions of predator-prey interactions. Am Nat 97: 209-223")
          ))

   
  )
)
)
)

# This creates the "behind the scenes" code (Server)
server = function(input, output){
RM=function(t, y, parameters){
  N=y[1]
  P=y[2]

  r=parameters["r"]
  K=parameters["K"]
  a=parameters["a"]
  c=parameters["c"]
  b=parameters["b"]
  g=parameters["g"]

    dN = r*N*(1-N/K)-a*N*P/(c+N)
    dP = b*N*P/(c+N)-g*P
    res=c(dN,dP)
    list(res)
}

output$plot1 <- renderPlot({

 times  = seq(0, input$Tmax, by=0.1)
 parms=c(r=input$r, K=input$K,a=input$a,
   c=input$c,b=input$b,g=input$g)
 xstart = c(N=input$N, P=input$P)

 out=ode(y=xstart,
    times=times,
    func=RM,
    parms=parms)

    out=as.data.frame(out)

  r=parms["r"]
  K=parms["K"]
  a=parms["a"]
  c=parms["c"]
  b=parms["b"]
  g=parms["g"]


 plot(out$time, out$N, ylab="abundance", xlab="time", type="l", ylim=range(out[,2:3]))
 lines(out$time, out$P, col="red")
    legend("topright",
          legend=c("N", "P"),
          lty=c(1,1),
           col=c("black", "red"))
})

output$plot2 <- renderPlot({

 times  = seq(0, input$Tmax, by=0.1)
 parms=c(r=input$r, K=input$K,a=input$a,
   c=input$c,b=input$b,g=input$g)
 xstart = c(N=input$N, P=input$P)

 out=ode(y=xstart,
    times=times,
    func=RM,
    parms=parms)

    out=as.data.frame(out)

  r=parms["r"]
  K=parms["K"]
  a=parms["a"]
  c=parms["c"]
  b=parms["b"]
  g=parms["g"]

#null clines
plot(out$N, out$P, ylab='predator', xlab='prey', type='l', 
xlim=range(out$N), ylim= range(out$P))
abline(h=0, col = "green")
abline(v=0, col = "red")
curve(r*(1-x/K)*(c+x)/a,from = 0, to = max(c(90, out$N)), col = "green",add=T)
abline(v=g*c/(b-g),col = "red")
fld=flowField(RM, xlim=range(out$N), ylim=range(out$P), 
parameters=parms, system="two.dim", add=TRUE)
   legend("topright",
          legend=c("N-iso", "P-iso"),
          lty=c(1,1),
           col=c("green", "red"))

# points(Nstar,Pstar,pch = 1)

})


}

shinyApp(ui, server, options = list(height = 500))
```

Reference:

Rosenzweig, M.L. and MacArthur, R.H. (1963) Graphical representation 
              and stability conditions of predator-prey interactions. Am Nat 97: 209-223
