---
title: "Ross-Macdonald model with embedded ShinyApp"
author: Ottar N. Bjørnstad
output: html_document
runtime: shiny
---

Version 0.5.3 August 26, 2019 
https://github.com/objornstad

This Rmarkdown of a Ross-Macdonald model (Smith et al. 2012) was written by Ottar N. Bjørnstad and is released with a CC-BY-NC license for anyone to improve and re-share (acknowledging origin). Please email me a copy of update (onb1 at psu dot edu).

The app requires the shiny and deSolve packages to be installed to run. 

```{r, echo=FALSE}
require(shiny)
require(deSolve)
#The flowField function was written by Michael J Grayling (https://github.com/mjg211) as part of 
#recently archived phaseR-package and is reproduced here under the original GPL-3 license
flowField=function (deriv, xlim, ylim, parameters = NULL, system = "two.dim", 
   points = 21, col = "gray", arrow.type = "equal", arrow.head = 0.05, 
    frac = 1, add = TRUE, xlab = "x", ylab = "y", state.names = c("x", 
        "y"), ...) 
{
    if ((!is.vector(xlim)) | (length(xlim) != 2)) {
        stop("xlim is not a vector of length 2 as required")
    }
    if (xlim[2] <= xlim[1]) {
        stop("xlim[2] is less than or equal to xlim[1]")
    }
    if ((!is.vector(ylim)) | (length(ylim) != 2)) {
        stop("ylim is not a vector of length 2 as required")
    }
    if (ylim[2] <= ylim[1]) {
        stop("ylim[2] is less than or equal to ylim[1]")
    }
    if (points <= 0) {
        stop("points is less than or equal to zero")
    }
    if (!(system %in% c("one.dim", "two.dim"))) {
        stop("system must either be set to \"one.dim\" or \"two.dim\"")
    }
    if (is.vector(col) == FALSE) {
        stop("col is not a vector as required")
    }
    if (length(col) > 1) {
        col <- col[1]
        print("Note: col has been reset as required")
    }
    if (!(arrow.type %in% c("proportional", "equal"))) {
        stop("arrow.type must either be set to \"proportional\" or \"equal\"")
    }
    if (arrow.head <= 0) {
        stop("arrow.head is less than or equal to zero")
    }
    if (frac <= 0) {
        stop("frac is less than or equal to zero")
    }
    if (!is.logical(add)) {
        stop("add must be logical")
    }
    x <- seq(from = xlim[1], to = xlim[2], length = points)
    y <- seq(from = ylim[1], to = ylim[2], length = points)
    dx <- matrix(0, ncol = points, nrow = points)
    dy <- matrix(0, ncol = points, nrow = points)
    xmax.length <- x[2] - x[1]
    ymax.length <- y[2] - y[1]
    if (add == FALSE) {
        plot(1, xlim = c(xlim[1] - xmax.length, xlim[2] + xmax.length), 
            ylim = c(ylim[1] - ymax.length, ylim[2] + ymax.length), 
            type = "n", xlab = xlab, ylab = ylab, ...)
    }
    if (system == "one.dim") {
        for (i in 1:points) {
            dy[1, i] <- deriv(0, setNames(c(y[i]), state.names[1]), 
                parameters)[[1]]
        }
        for (i in 2:points) {
            dy[i, ] <- dy[1, ]
        }
        abs.dy <- abs(dy)
        abs.dy.non <- abs.dy[which(abs.dy != 0)]
        max.abs.dy <- max(abs(dy))
        coefficient <- frac * min(xmax.length, ymax.length)/(2 * 
            sqrt(2) * max(sqrt(2 * abs.dy.non/(abs.dy.non + (1/abs.dy.non))), 
            sqrt(2 * (1/abs.dy.non)/(abs.dy.non + (1/abs.dy.non)))))
        for (i in 1:points) {
            for (j in 1:points) {
                if (dy[i, j] != 0) {
                  factor <- sqrt(2/(abs.dy[i, j] + (1/abs.dy[i, 
                    j])))
                  y.shift <- coefficient * factor * sqrt(abs.dy[i, 
                    j])
                  x.shift <- coefficient * factor/sqrt(abs.dy[i, 
                    j])
                  if (dy[i, j] < 0) {
                    y.shift <- -y.shift
                  }
                }
                if (dy[i, j] == 0) {
                  y.shift <- 0
                  x.shift <- coefficient * sqrt(2)
                }
                if (arrow.type == "proportional") {
                  if (dy[i, j] != 0) {
                    prop <- abs.dy[i, j]/max.abs.dy
                    y.shift <- y.shift * prop
                    x.shift <- x.shift * prop
                  }
                  if (dy[i, j] == 0) {
                    x.shift <- y.shift * mean(abs.dy)/max.abs.dy
                  }
                }
                arrows(x[i] - x.shift, y[j] - y.shift, x[i] + 
                  x.shift, y[j] + y.shift, length = arrow.head, 
                  col = col, ...)
            }
        }
        return(list(add = add, arrow.head = arrow.head, arrow.type = arrow.type, 
            col = col, deriv = deriv, dy = dy, frac = frac, parameters = parameters, 
            points = points, system = system, x = x, xlab = xlab, 
            xlim = xlim, y = y, ylab = ylab, ylim = ylim))
    }
    else {
        for (i in 1:length(x)) {
            for (j in 1:length(y)) {
                df <- deriv(0, setNames(c(x[i], y[j]), state.names), 
                  parameters)
                dx[i, j] <- df[[1]][1]
                dy[i, j] <- df[[1]][2]
            }
        }
        abs.dx <- abs(dx)
        abs.dy <- abs(dy)
        abs.dx.non <- abs.dx[which((abs.dx != 0) & (abs.dy != 
            0))]
        abs.dy.non <- abs.dy[which((abs.dx != 0) & (abs.dy != 
            0))]
        max.length <- max(sqrt(dx^2 + dy^2))
        coefficient <- frac * min(xmax.length, ymax.length)/(2 * 
            sqrt(2) * max(sqrt(2 * (abs.dy.non/abs.dx.non)/((abs.dy.non/abs.dx.non) + 
            (abs.dx.non/abs.dy.non))), sqrt(2 * (abs.dx.non/abs.dy.non)/((abs.dy.non/abs.dx.non) + 
            (abs.dx.non/abs.dy.non)))))
        for (i in 1:points) {
            for (j in 1:points) {
                if ((dx[i, j] != 0) | (dy[i, j] != 0)) {
                  if ((dx[i, j] != 0) & (dy[i, j] != 0)) {
                    factor <- sqrt(2/((abs.dy[i, j]/abs.dx[i, 
                      j]) + (abs.dx[i, j]/abs.dy[i, j])))
                    y.shift <- coefficient * factor * sqrt(abs.dy[i, 
                      j]/abs.dx[i, j])
                    x.shift <- coefficient * factor/sqrt(abs.dy[i, 
                      j]/abs.dx[i, j])
                    if (dy[i, j] < 0) {
                      y.shift <- -abs(y.shift)
                    }
                    if (dx[i, j] < 0) {
                      x.shift <- -abs(x.shift)
                    }
                  }
                  if ((dx[i, j] == 0) & (dy[i, j] != 0)) {
                    y.shift <- coefficient * sqrt(2)
                    x.shift <- 0
                    if (dy[i, j] < 0) {
                      y.shift <- -abs(y.shift)
                    }
                  }
                  if ((dx[i, j] != 0) & (dy[i, j] == 0)) {
                    y.shift <- 0
                    x.shift <- coefficient * sqrt(2)
                    if (dx[i, j] < 0) {
                      x.shift <- -abs(x.shift)
                    }
                  }
                  if (arrow.type == "proportional") {
                    prop <- sqrt((abs.dx[i, j]^2 + abs.dy[i, 
                      j]^2))/max.length
                    y.shift <- y.shift * prop
                    x.shift <- x.shift * prop
                  }
                  arrows(x[i] - x.shift, y[j] - y.shift, x[i] + 
                    x.shift, y[j] + y.shift, length = arrow.head, 
                    col = col, ...)
                }
            }
        }
    }
    return(list(add = add, arrow.head = arrow.head, arrow.type = arrow.type, 
        col = col, deriv = deriv, dx = dx, dy = dy, frac = frac, 
        parameters = parameters, points = points, system = system, 
        x = x, xlab = xlab, xlim = xlim, y = y, ylab = ylab, 
        ylim = ylim))
}

```

The basic equations due to Aaron and May (1982) are:
$$
dx/dt = (a b Y / X) y (1-x) -\gamma x \\
dy/dt = a c x (1-y)-  \mu y,
$$
where $x$ is the fraction of infected humans, $y$ is fraction of infected mosquitos, m = Y/X is mosquito-to-human ratio, $\gamma$ is human recovery rate, $1/\mu$ is mosquito life expectancy, $a$ is biting rate (1 / gonotrophic-cycle), $b$ is human probability of getting infected by infected mosquito, $c$ is probability of mosquito infection from infected human.

The isoclines (sometimes called the nullclines) in this system are given by the solution to the 
equations $dx/dt=0$ and $dy/dt=0$ and partitions the phase plane into regions 
were $x$ and $y$ are increasing and decreasing. 
$$
x^*=\frac{\gamma x}{(a b m)(1-x)}\\
y^* =a c x / (a c x + \mu)
$$
The basic reproductive ratio ($R_0$) and endemic equilibrium ($\{x^*, y^*\}$) are:
$$
R_0 = m a^2 b c / \mu \gamma\\
x^*=(R_0-1)/[R_0 + a c / \mu]\\
y^*=(\frac{R_0 - 1}{R_0})(\frac{ac/\mu}{1+ac/\mu})
$$



The shiny app:

```{r, echo=FALSE}
# This creates the User Interface (UI)
ui <- pageWithSidebar(
headerPanel(""),
sidebarPanel(
sliderInput("gamma", "gamma", 1/7,
              min = 0, max = 1),
sliderInput("a", "a", 1/4,
              min = 0, max = 1),
sliderInput("b", "b", 0.5,
              min = 0, max = 1),
sliderInput("c", "c", 0.9,
              min = 0, max = 1),
sliderInput("mu", "mu", 1/7,
              min = 0, max = 1),
sliderInput("m", "m", 10,
              min = 1, max = 100)
),
mainPanel(
  plotOutput("plot1", height = 500)
)
)


# This creates the 'behind the scenes' code (Server)
server <- function(input, output) {
grfn=function(t, y, parameters){
  X=y[1]
  Y=y[2]
  with(as.list(parameters),{
  dx=a*b*m*Y*(1-X)-gamma*X
  dy=a*c*X*(1-Y)-mu*Y
  gr=c(dx, dy)
  list(gr)
  })
}


  output$plot1 <- renderPlot({
times=seq(0, 365*2, by=.1)

parameters  = c(gamma = input$gamma, a =  input$a, b=input$b, c=input$c, mu=input$mu, m=input$m)
start=c(0.01, 0.01)

out=ode(y=start,
  times=times,
  func=grfn,
  parms=parameters)

  out=as.data.frame(out)

with(as.list(parameters),{
curve(gamma*x/((a*b*m)*(1-x)), 0,1, ylim=c(0,1), xlab="x", ylab="y")
R0=m*a^2*b*c/(mu*gamma)
title(paste ("R0=",round(R0,2)))
curve(a*c*x/(a*c*x+mu), 0,1, add=TRUE, col="red")
fld=flowField(grfn, xlim=c(0,1), ylim=c(0,1), 
parameters=parameters, system="two.dim", add=TRUE,
ylab="H", xlab="M")

})
points(out[,2], out[, 3])
legend("topleft", c("H isocline", "M isocline", "Trajectory"), lty=c(1,1,0), col=c(1,2, 1), pch=c(NA,NA, 1))
   })
  }

shinyApp(ui, server, options = list(height = 680))
```

References:
Aron, J.L. and May, R.M. 1982. The population dynamics of malaria. In The population dynamics of infectious diseases: theory and applications (pp. 139-179). Springer, Boston, MA.

Smith, D.L., Battle, K.E., Hay, S.I., Barker, C.M., Scott, T.W. and McKenzie, F.E. 2012. Ross, Macdonald, and a theory for the dynamics and control of mosquito-transmitted pathogens. PLoS pathogens, 8(4), p.e1002588.
