---
title: "SIR model with ShinyApp"
author: Ottar N. Bjørnstad
output: html_document
runtime: shiny
---

Version 0.5.2 January 23, 2020
https://github.com/objornstad

This Rmarkdown of the SIR model was written by Ottar N. Bjørnstad and is released with a CC-BY-NC lisence for anyone to improve and re-share (acknowledging origin). Please email me a copy of update (onb1 at psu dot edu). The app was originally developed as part of the epimdr-package (https://cran.r-project.org/package=epimdr; Bjørnstad 2019).

The app requires the shiny and deSolve packages to be installed to run.

```{r, echo=FALSE}
require(shiny)
require(deSolve)
require(phaseR)
```

The basic equations for the flow of hosts between **S**usceptible, **I**nfectious and **R**ecovered
 compartments are:
\begin{align}
    \frac{dS}{dt} =& \underbrace{\mu N}_{\mbox{birth}} - \underbrace{\beta I \frac{S}{N}}_{\mbox{infection}} - \underbrace{\mu S}_{\mbox{death}} \label{eq:sirs}\\
     \frac{dI}{dt} =& \underbrace{\beta I \frac{S}{N}}_{\mbox{infection}} - \underbrace{\gamma I}_{\mbox{recovery}} - \underbrace{\mu I}_{\mbox{death}}  \label{eq:siri}\\
     \frac{dR}{dt} =& \underbrace{\gamma I}_{\mbox{recovery}} - \underbrace{\mu R}_{\mbox{death}} \label{eq:sirr}
\end{align}
The assumptions of this version of the SIR model are:

 
- The infection circulates 
     in a population of size $N$, with a per capita ``background"
     death rate, $\mu$, which is balanced by a
     birth rate $\mu N$. From the sum of the equations $dN/dt=0$ and $N=S+I+R$ is constant.

- The infection causes morbidity (not mortality).

- Newborns are recruited directly into the susceptible
     class at birth.

- Transmission of infection from infectious to
     susceptible individuals is controlled by a bilinear contact
     term $\beta I \frac{S}{N}$, from the assumption that the $I$ infectious individuals are independently and randomly
     mixing with all other individuals, so the fraction $S/N$ is with susceptible individuals; $\beta$ is the transmission rate.

- Infected individuals move directly into the 
     the infectious class and remains there 
     for an average infectious period of $1/(\gamma+\mu)$ time units.

- Recovered individuals are immune
     from re-infection for life.

For the basic SIR model $R_0 = \frac{\beta}{\gamma + \mu}$. The isoclines (sometimes called the null-clines)  are given by the solution to the 
equations $dS/dt=0$ and $dI/dt=0$ and partitions the phase plane into regions 
were $S$ and $I$ are increasing and decreasing. 
For $N=1$, the $I$-isocline is $S = (\gamma +\mu)/\beta = 1/R_0$
and the S-isocline is $I= \mu (1/S-1)/\beta$.

The resonant period is $\frac{2 \pi}{\mu (R_0 -1) (\mu + \gamma)}$ 

The  shiny app:

```{r, echo=FALSE}
# This creates the User Interface (UI)
ui <- pageWithSidebar(
headerPanel(""),
#The sidebar for parameter input
sidebarPanel(
#Sliders:
sliderInput("beta", "Transmission (yr^-1):", 300,
              min = 0, max = 1000),
sliderInput("infper", "Infectious period (days)", 5,
              min = 1, max = 100),
sliderInput("mu", "birth rate:", 5,
              min = 0, max = 100),
sliderInput("T", "Time range:",
                  min = 0, max = 1, value = c(0,1)),
width=3
),
#Main panel for figures and equations
mainPanel(
  #Multiple tabs in main panel
  tabsetPanel(
      #Tab 1: Time plot (plot1 from server)
      tabPanel("Time", plotOutput("plot1")), 
      #Tab 2: Phase plot (plot2 from server)
      tabPanel("Phase plane", plotOutput("plot2", height = 500)),
      #Tab 3: MathJax typeset equations 
      tabPanel("Equations", 
           withMathJax(
            helpText("Susceptible $$\\frac{dS}{dt} = \\mu (N - S) - \\frac{\\beta I S}{N}$$"),
            helpText("Infecitous $$\\frac{dI}{dt} = \\frac{\\beta I S}{N} - (\\mu+\\sigma) I$$"),
           helpText("Removed $$\\frac{dR}{dt} = \\gamma I - \\mu R$$"),
           helpText("Reproductive ratio $$R_0 =  \\frac{1}{\\gamma+\\mu} \\frac{\\beta N}{N}$$")             
           ))
  ))) #End of ui()


# This creates the 'behind the scenes' code (Server)
server <- function(input, output) {
  #Gradient function for SIR model
  sirmod=function(t, x, parms){
    S=x[1]
    I=x[2]
    R=x[3]
    beta=parms["beta"]
    mu=parms["mu"]
    gamma=parms["gamma"]
    N=parms["N"]
    dS = mu * (N  - S)  - beta * S * I / N
    dI = beta * S * I / N - (mu + gamma) * I
    dR = gamma * I - mu * R
    res=c(dS, dI, dR)
    list(res)
  }

 #Plot1: renderPlot to be passed to UI tab 1
  output$plot1 <- renderPlot({
  #input\$xx's are pulled from UI
  times  = seq(0, input$T[2], by=1/1000)
  parms  = c(mu = input$mu, N = 1, beta =  input$beta, gamma =
    365/input$infper)
  start = c(S=0.999, I=0.001, R = 0)
  R0 = round(with(as.list(parms), beta/(gamma+mu)), 1)
  #Resonant period
  AA=with(as.list(parms), 1/(mu*(R0-1)))
  GG=with(as.list(parms), 1/(mu+gamma))
  rp=round(2*pi*sqrt(AA*GG),2)

  #Integrate ode with parameters pulled from UI
  out=ode(y=start,  times=times, func=sirmod, parms=parms)
  out=as.data.frame(out)

  #Plot1
  sel=out$time>input$T[1]&out$time<input$T[2]
  plot(x=out$time[sel], y=out$S[sel], ylab="fraction", xlab="time", type="l",
  ylim=range(out[sel,-c(1,4)]))
  title(paste("R0=", R0, "Period=", rp, "/ year"))
  lines(x=out$time[sel], y=out$I[sel], col="red")
  lines(x=out$time[sel], y=out$R[sel], col="green")
  legend("right",
        legend=c("S", "I", "R"),
        lty=c(1,1,1),
         col=c("black", "red", "green"))
   })

 #Plot2: renderPlot to be passed to UI tab 2
  output$plot2 <- renderPlot({
  times  = seq(0, input$T[2], by=1/1000)
  parms  = c(mu = input$mu, N = 1, beta =  input$beta, gamma =
    365/input$infper)
  start = c(S=0.999, I=0.001, R = 0)
  R0 = round(with(as.list(parms), beta/(gamma+mu)), 1)

  #Gradient function used for phaseR phase-plot
  simod=function(t, y, parameters){
   S=y[1]
   I=y[2]
   beta=parameters["beta"]
   mu=parameters["mu"]
   gamma=parameters["gamma"]
   N=parameters["N"]   
   dS = mu * (N  - S)  - beta * S * I / N
   dI = beta * S * I / N - (mu + gamma) * I
   res=c(dS, dI)
   list(res)
  }

  #Integrate simod
  out=ode(y=start[-3], times=times, func=simod, parms=parms)
  out=as.data.frame(out)

  AA=with(as.list(parms), 1/(mu*(R0-1)))
  GG=with(as.list(parms), 1/(mu+gamma))
  rp=round(2*pi*sqrt(AA*GG),2)
  
  plot(x=out$S, y=out$I, xlab="Fraction suceptible", ylab="Fraction infected", type="l")
  title(paste("R0=", R0, "Period=", rp, "/year"))
  #Add vector field
  fld=flowField(simod, xlim=range(out$S), ylim=range(out$I), 
  parameters=parms, system="two.dim", add=TRUE,
  ylab="I", xlab="S")
  #Add isoclines
  abline(v=1/R0, col="green")
  curve(parms["mu"]*(1-x)/(parms["beta"]*x), min(out$S), max(out$S), add=TRUE, col="red")
    legend("topright",
        legend=c("S-isocline", "I-isocline"),
        lty=c(1,1),
         col=c("red", "green"))
   })
  } #End of server()

shinyApp(ui, server, options = list(height = 500))
```

References:

Bjørnstad, O.N. (2018) Epidemics: Models and Data using R. Springer (312 pp) ISBN 978-3-319-97487-3 https://www.springer.com/gp/book/9783319974866



