---
title: "Parasitoid-host Coupled Map Lattice model"
author: "Ottar N. Bjornstad"
date: "8/21/2019"
output: html_document
runtime: shiny
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

Version 0.5.3 August 31, 2019 
https://github.com/objornstad/

This Rmarkdown of the Spatially-extended parasitoid-host Model was written by Ottar N. Bjørnstad and is released with a CC-BY-NC lisence for anyone to improve and re-share (acknowledging origin). Please email me a copy of update (onb1 at psu dot edu). The app was originally developed as part of the "epimdr-package for the "Epidemics: Models and Data UseR book (https://cran.r-project.org/package=epimdr; Bjørnstad 2019).

The app requires the shiny and plotly packages to be installed to run.

The Nicholson-Bailey (1935) assumes random search by parasitoids (with a searching efficiency $a$) by the parasitoids. The probability of escaping parasitation is $\exp(-a P_t)$ and the number of host, $H$, and parasitoids, $P$, in the next generation is:
\begin{eqnarray}
&H_{t+1} & = R H_t \exp(-a P_t)\\
&P_{t+1} & = R H_t (1-\exp(-a P_t)),
\end{eqnarray}
where $R$ is the average number of offspring per hosts. 

Hassell et al. (1991; see also Bjørnstad and Bascompte 2001) introduced the spatially-extended version of the model which asssumes a lattice landscape where a fraction of hosts and parasitoids disperse to neighboring patches between each generation. The result is a range of emergent spatiotemporal dynamics including spiral waves, spatial chaos and Turing crystals. 

```{r, echo=FALSE, message=FALSE}
require(shiny)
require(plotly)


# Define UI for slider demo app ----
ui <- fluidPage(

  # App title ----
  titlePanel("Sliders"),

  # Sidebar layout with input and output definitions ----
  sidebarLayout(

    # Sidebar to demonstrate various slider options ----
    sidebarPanel(

  
      sliderInput("Dh", "Host dispersal:", 0.5,
              min = 0, max = 1),
      sliderInput("Dp", "Parasitoid dispersal:", 0.5,
              min = 0, max = 1),
      sliderInput("R", "Host growth:", 2,
              min = 1, max = 5, step=0.1),
      sliderInput("a", "Parasitoid efficiency:", 1,
              min = 0, max = 5),
      numericInput("IT", "Generations:", 100,
               min = 0, max = 1000)
    ),

    # Main panel for displaying outputs ----
 mainPanel(tabsetPanel(
  tabPanel("Simulation", plotlyOutput("plot1", height = 500)),
     tabPanel("Details", 
     withMathJax(
                helpText("MODEL:"),
          helpText("Host $$ $$"),
          helpText("Parasitoid $$ $$")))
           )
)
)
)
# Define server logic for slider examples ----
server <- function(input, output) {

#xlen is width of the lattice (E-W)
#ylen is height of the lattice (N-S)
xlen = 30
ylen = 30

hp.dyn = function(h, p, R, a){ 
   #hnew is the post-interaction host density
   hnew = R * h * exp(- a * p)
   #pnew is the post-interaction parasitoid density
   pnew = R * h * (1 - exp(- a * p))
   #the two vectors of results are stored in a "list"
   res = list(h = hnew, p = pnew)
   return(res)
} 

xy = expand.grid(1:xlen, 1:ylen)
dmat = as.matrix(dist(xy))



  # Show the values in an HTML table ----
output$plot1 <- renderPlotly({
kh = ifelse(dmat<1.5,input$Dh/8,0)
kp = ifelse(dmat<1.5,input$Dp/8,0)
diag(kh) = 1-input$Dh
diag(kp) = 1-input$Dp

IT = input$IT
hmat = matrix(NA, nrow=xlen*ylen, ncol = IT)
pmat = matrix(NA, nrow=xlen*ylen, ncol = IT)
hmat[,1] = 0
pmat[,1] = 0
hmat[23,1] = 4
pmat[23,1] = 1

for(i in 2:IT){
   #growth
   tmp = hp.dyn(h = hmat[,(i-1)], p = pmat[,(i-1)], 
      R = input$R, a = input$a)
   #redistribution
   hmat[,i] = tmp$h%*%kh;
   pmat[,i] = tmp$p%*%kp;
}

hmat2=as.data.frame(hmat)
hmat2$x=xy[,1]
hmat2$y=xy[,2]
longH=reshape(hmat2, direction="long", varying=1:IT, v.names="H")

require(plotly)
anim=ggplot(longH, aes(x=x, y=y, frame=longH$time))+geom_point(size=longH$H)
ggplotly(anim)
})

}

# Create Shiny app ----
shinyApp(ui, server, options = list(height = 750))
```


REFERENCES:

Bjørnstad, O. N., & Bascompte, J. (2001). Synchrony and second-order spatial correlation
in host–parasitoid systems. Journal of Animal Ecology, 70(6), 924–933.

Hassell, M. P., Comins, H. N., & May, R. M. (1991). Spatial structure and chaos in
insect population dynamics. Nature, 353(6341), 255–258.

Nicholson, A. J., & Bailey, V. A. (1935). The balance of animal populations. Part I.
Proceedings of the Zoological Society of London, 105(3), 551–598.

